---
title:   "Sex Variation in the DGRP Metabolome - Univariate means analysis"
output:   html_notebook
by:       R. Coig
updated:  5/7/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Univariate analysis, TAG-Metabolite level correlations (Sample Means, without sex as a covariate)
```{r,load required libraries}
library(tidyr)#Data manipulation
library(dplyr)#Data manipulation
library(smatr)#Computing relative TAG
library(reshape2)
```
Loading in metabolite means data 
```{r,loading in metabolome data}
rm(list = ls())
load('../data/processed/mz_means.data')
load('../data/processed/batch_corrected.data')
load('../data/processed/phenotype-TAG.data')
load('../data/processed/phenotype-mass.data')
```
TAG Correlations across all metabolites for sex/geno means
```{r}
#formatting df
mz_means_by_TAG<-merge(mz_means_by_sex_geno,TAG_sex_geno_long,by=c("sex","genotype"))

# Create an empty results data frame to store the results
TAG_mz_COR_means_results <- data.frame(COMPOUND = character(0), 
                                    TAG_R_spearman = numeric(0),
                                    TAG_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(mz_means_by_TAG, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$tTAG_4_mu_ss, subset_data$mean, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          TAG_R_spearman = spearman_result$estimate,
                          TAG_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    TAG_mz_COR_means_results <- distinct(rbind(TAG_mz_COR_means_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
TAG_mz_COR_means_results$TAG_adj <- p.adjust(TAG_mz_COR_means_results$TAG_p, method = "fdr")
TAG_mz_COR_means_results<-TAG_mz_COR_means_results%>%arrange(TAG_p)
TAG_mz_COR_means_results$sig_TAG <- ifelse(TAG_mz_COR_means_results$TAG_adj < 0.01, "SIG", "NS")

sig_meanTAG_mz<-TAG_mz_COR_means_results%>%filter(sig_TAG=="SIG")%>%pull(COMPOUND)
```
Mass Correlations across all metabolites for sex/geno means
```{r}
#formatting df
mz_means_by_mass<-merge(mz_means_by_sex_geno,mass_sex_geno_long,by=c("sex","genotype"))
#Removing sex/geno combos that do not have mass values
mz_means_by_mass<-mz_means_by_mass%>%filter(!is.na(mass_4_mu_ss))

# Create an empty results data frame to store the results
mass_mz_COR_means_results <- data.frame(COMPOUND = character(0), 
                                    mass_R_spearman = numeric(0),
                                    mass_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(mz_means_by_mass, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$mass_4_mu_ss, subset_data$mean, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          mass_R_spearman = spearman_result$estimate,
                          mass_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    mass_mz_COR_means_results <- distinct(rbind(mass_mz_COR_means_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
mass_mz_COR_means_results$mass_adj <- p.adjust(mass_mz_COR_means_results$mass_p, method = "fdr")
mass_mz_COR_means_results<-mass_mz_COR_means_results%>%arrange(mass_p)
mass_mz_COR_means_results$sig_mass <- ifelse(mass_mz_COR_means_results$mass_adj < 0.01, "SIG", "NS")

sig_meanmass_mz<-mass_mz_COR_means_results%>%filter(sig_mass=="SIG")%>%pull(COMPOUND)
```
Creating relative TAG as the residuals of TAG~Mass
```{r}
rTAG_sex_geno_long<-merge(mass_sex_geno_long,TAG_sex_geno_long,by=c("sex","genotype"))%>%select(sex,genotype,tTAG_4_mu_ss,mass_4_mu_ss)%>%filter(!is.na(mass_4_mu_ss))
rTAG_sex_geno_long<-rTAG_sex_geno_long%>%filter(!is.na(tTAG_4_mu_ss))

# Perform SMA regression
sma_model <- sma(rTAG_sex_geno_long$tTAG_4_mu_ss ~ rTAG_sex_geno_long$mass_4_mu_ss, na.rm=T)

# Extract the residuals
rTAG_sex_geno_long$rTAG_4_mu_ss <- residuals(sma_model)

mz_means_by_rTAG<-merge(mz_means_by_sex_geno,rTAG_sex_geno_long,by=c("sex","genotype"))

```
rTAG Correlations across all metabolites for means
```{r}
# Create an empty results data frame to store the results
rTAG_mz_COR_means_results <- data.frame(COMPOUND = character(0), 
                                    rTAG_R_spearman = numeric(0),
                                    rTAG_p = numeric(0),
                                    SSD = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(mz_means_by_rTAG, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$rTAG_4_mu_ss, subset_data$mean, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          rTAG_R_spearman = spearman_result$estimate,
                          rTAG_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    rTAG_mz_COR_means_results <- distinct(rbind(rTAG_mz_COR_means_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
rTAG_mz_COR_means_results$rTAG_adj <- p.adjust(rTAG_mz_COR_means_results$rTAG_p, method = "fdr")
rTAG_mz_COR_means_results<-rTAG_mz_COR_means_results%>%arrange(rTAG_p)
rTAG_mz_COR_means_results$sig_rTAG <- ifelse(rTAG_mz_COR_means_results$rTAG_adj < 0.01, "SIG", "NS")

sig_meanrTAG_mz<-rTAG_mz_COR_means_results%>%filter(sig_rTAG=="SIG")%>%pull(COMPOUND)
```
Combining Phenotype Results
```{r}
Pheno_Results_Means<-merge(TAG_mz_COR_means_results,mass_mz_COR_means_results,by="COMPOUND")
Pheno_Results_Means<-merge(Pheno_Results_Means,rTAG_mz_COR_means_results,by="COMPOUND")
```
Saving out results for summary document
```{r}
write.csv(TAG_mz_COR_means_results,"output/03_Mean_tag_mz_results.csv",row.names=F)
write.csv(mass_mz_COR_means_results,"output/03_Mean_mass_mz_results.csv",row.names=F)
write.csv(rTAG_mz_COR_means_results,"output/03_Mean_rtag_mz_results.csv",row.names=F)
write.csv(Pheno_Results_Means,"output/03_Mean_pheno_mz_results.csv",row.names=F)
```
Saving out r files
```{r}
save(file='../data/results/TAG_mz_Means_COR_results.data', Pheno_Results_Means, sig_meanTAG_mz, sig_meanmass_mz, mz)
rm(list = ls())
```