---
title:   "Sex Variation in the DGRP Metabolome - Univariate analysis"
output:   html_notebook
by:       R. Coig
updated:  3/31/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Correlations for deltaTAG by sex differences in metabolite level
```{r,load required libraries}
#library(tidyr)#Data manipulation
library(dplyr)#Data manipulation
library(car)#For specificying Type III ANOVA
```
Loading in batch corrected data for analysis
```{r,loading in batch corrected metabolome data}
rm(list = ls())
load('../data/processed/mz_SD.data')
load('../data/processed/phenotype-dimorphism.data')
SD_by_geno<-merge(SD_by_geno,dimorphism_summary,by="genotype")
```
Spearman's Correlation with deltaTAG and delta mean metabolite levels
```{r}
# Create an empty results data frame to store the results
deltaTAG_SDmz_COR_results <- data.frame(COMPOUND = character(0), 
                                    deltaTAG_R_spearman = numeric(0),
                                    deltaTAG_p = numeric(0),
                                    SSD = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(SD_by_geno, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$mz_SD, subset_data$diff_TAG, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          deltaTAG_R_spearman = spearman_result$estimate,
                          deltaTAG_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    deltaTAG_SDmz_COR_results <- distinct(rbind(deltaTAG_SDmz_COR_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
deltaTAG_SDmz_COR_results$deltaTAG_adj <- p.adjust(deltaTAG_SDmz_COR_results$deltaTAG_p, method = "fdr")
deltaTAG_SDmz_COR_results<-deltaTAG_SDmz_COR_results%>%arrange(deltaTAG_p)
deltaTAG_SDmz_COR_results$sig_deltaTAG <- ifelse(deltaTAG_SDmz_COR_results$deltaTAG_adj < 0.01, "SIG", "NS")

sig_SD_deltaTAG<-deltaTAG_SDmz_COR_results%>%filter(sig_deltaTAG=="SIG")%>%pull(COMPOUND)
```
Saving out Univariate results for deltaTAG correlations
```{r}
write.csv(deltaTAG_SDmz_COR_results,"output/03_deltaTAG_SDmz_COR_results.csv",row.names=F)
```
Saving out r files
```{r}
save(file='../data/results/deltaTAG_SDmz_COR_results.data', deltaTAG_SDmz_COR_results, sig_SD_deltaTAG, mz)
rm(list = ls())
```