---
title:   "Sex Variation in the DGRP Metabolome - Univariate Analysis"
output:   html_notebook
by:       R. Coig
updated:  5/7/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary:  Univariate summary of metabolite levels
```{r,load required libraries}
library(dplyr)
library(tidyr)
library(reshape2)
```
Loading in batch corrected data for analysis
```{r,loading in batch corrected metabolome data}
rm(list = ls())
load('../data/processed/batch_corrected.data')
```
Making all values positive for calculating SD (F-M) and log SD log(F/M)
```{r,positive values}
melted_dat_sd<-melted_dat
minvalue<-min(melted_dat_sd$value, na.rm=T)
melted_dat_sd$value<-(melted_dat_sd$value-minvalue+1)#making all values positive
```
Calculating the sex difference in metabolite level for each genotype by subtracting values
```{r}
# Create an empty list to store the results
SD_by_geno <- list()
# Loop through each unique COMPOUND in mz
for (compound in mz) {
  # Filter the data for the current COMPOUND
  filtered_data <- melted_dat_sd %>% filter(COMPOUND == compound)
  # Calculate the difference between female and male mz levels
  diff_geno <- filtered_data %>%
    dplyr::group_by(genotype, sex) %>%
    dplyr::reframe(mean = mean(value, na.rm = TRUE)) %>%
    tidyr::pivot_wider(names_from = sex, values_from = c(mean))%>%
    #dplyr::mutate(mz_diff = mean_female-mean_male) %>%
    dplyr::reframe(genotype, 
                     COMPOUND = compound,
                     mz_SD = (female - male),
                     mz_logSD = (log(female/male)))
  # Add the result to the list
  SD_by_geno[[compound]] <- diff_geno
}
SD_by_geno <- bind_rows(SD_by_geno)
rm(diff_geno)
```
Calculating the means by sex and genotype for each group using original melted data
```{r}
# Create an empty list to store the results
mz_means_by_sex_geno <- list()
# Loop through each unique COMPOUND in mz
for (compound in mz) {
  # Filter the data for the current COMPOUND
  filtered_data <- melted_dat %>% filter(COMPOUND == compound)
  # Calculate summaries for the current COMPOUND, grouping by genotype and sex
  summary <- filtered_data %>%
    dplyr::group_by(genotype, sex) %>%
    dplyr::reframe(COMPOUND=COMPOUND,
                   mean=mean(value,na.rm=TRUE),
                   variance = var(value, na.rm = TRUE),
                   sd = sd(value, na.rm = TRUE),
                   range = max(value) - min(value),
                   se = sd(value, na.rm = TRUE) / sqrt(n()),  # Calculate standard error
                   n_sex_geno = n()
)
  # Add the result to the list
  mz_means_by_sex_geno[[compound]] <- summary
}
mz_means_by_sex_geno <- bind_rows(mz_means_by_sex_geno)%>%distinct()
rm(filtered_data,summary)
```
Calculating the means by sex for each group using original melted data
```{r}
# Create an empty list to store the results
mz_means_by_sex <- list()
# Loop through each unique COMPOUND in mz
for (compound in mz) {
  # Filter the data for the current COMPOUND
  filtered_data <- melted_dat %>% filter(COMPOUND == compound)
  # Calculate the summaries for each COMPOUND, grouping by sex
  summary <- filtered_data %>%
    dplyr::group_by(sex) %>%
    dplyr::reframe(COMPOUND=COMPOUND,
                   mean=mean(value,na.rm=TRUE),
                   variance = var(value, na.rm = TRUE),
                   sd = sd(value, na.rm = TRUE),
                   range = max(value) - min(value),
                   se = sd(value, na.rm = TRUE) / sqrt(n()),  # Calculate standard error
                   n_sex = n()
)
  # Add the result to the list
  mz_means_by_sex[[compound]] <- summary
}
mz_means_by_sex <- bind_rows(mz_means_by_sex)%>%distinct()
rm(filtered_data,summary)
```
Calculating the means by genotype for each group using original melted data
```{r}
# Create an empty list to store the results
mz_means_by_geno <- list()
# Loop through each unique COMPOUND in mz
for (compound in mz) {
  # Filter the data for the current COMPOUND
  filtered_data <- melted_dat %>% filter(COMPOUND == compound)
  # Calculate the summaries for the current COMPOUND, grouping by genotype
  summary <- filtered_data %>%
    dplyr::group_by(genotype) %>%
    dplyr::reframe(COMPOUND=COMPOUND,
                   mean=mean(value,na.rm=TRUE),
                   variance = var(value, na.rm = TRUE),
                   sd = sd(value, na.rm = TRUE),
                   range = max(value) - min(value),
                   se = sd(value, na.rm = TRUE) / sqrt(n()),  # Calculate standard error
                   n_samples = n()
)
  # Add the result to the list
  mz_means_by_geno[[compound]] <- summary
}
mz_means_by_geno <- bind_rows(mz_means_by_geno)%>%distinct()
rm(filtered_data,summary)
```
Calculating the Euclidean distance across metabolites for each genotype *Not sure if this is correct??
```{r}
# Spread the data from long to wide format
mz_means_by_sex_geno_wide <- mz_means_by_sex_geno[,c(1:4)]
mz_means_by_sex_geno_wide <- pivot_wider(data = mz_means_by_sex_geno_wide, names_from = COMPOUND, values_from = mean)

female_means<-mz_means_by_sex_geno_wide%>%filter(sex=="female")%>%select(-genotype,-sex)
male_means<-mz_means_by_sex_geno_wide%>%filter(sex=="male")%>%select(-genotype,-sex)

# Number of genotypes
n_genotypes <- nrow(female_means)

# Initialize a vector to store the Euclidean distances for each genotype
ESD <- numeric(n_genotypes)

# Calculate Euclidean distance for each genotype
for(i in 1:n_genotypes) {
  ESD[i] <- sqrt(sum((female_means[i, ] - male_means[i, ])^2))
}

euclidean_distances<-as.data.frame(ESD)
euclidean_distances$genotype<-levels(mz_means_by_sex_geno$genotype)

SD_by_geno<-merge(SD_by_geno,euclidean_distances,by="genotype")
```
Calculating number of samples per genotype for weighting regressions
```{r}
n_samples <- key %>%
  dplyr::group_by(genotype) %>%
  dplyr::summarise(n_samples = dplyr::n())

mz_means_by_sex_geno<-merge(mz_means_by_sex_geno,n_samples,by="genotype")
```
Saving out summaries
```{r}
write.csv(SD_by_geno,"output/01_Metabolite_within_genotype_SD.csv",row.names=F)

write.csv(mz_means_by_sex_geno,"output/01_Metabolite_means_by_sex_and_genotype.csv",row.names=F)
write.csv(mz_means_by_sex,"output/01_Metabolite_means_by_sex.csv",row.names=F)
write.csv(mz_means_by_geno,"output/01_Metabolite_means_by_genotype.csv",row.names=F)
```
Saving out r files
```{r}
save(file='../data/processed/mz_means.data', mz_means_by_sex_geno, mz_means_by_sex, mz_means_by_geno, mz)
save(file='../data/processed/mz_SD.data', SD_by_geno, mz)
rm(list = ls())
```