---
title:   "Sex Variation in the DGRP Metabolome - AIC"
output:   html_notebook
by:       R. Coig
updated:  5/7/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary:  AIC analysis for sex, genotype and interaction across PCs
```{r,load required libraries}
library(dplyr)
library(tidyr)
```
Loading in PCA data for analysis
```{r,loading in batch corrected metabolome data}
rm(list = ls())
load('../data/processed/PCA.data')
```
Calculating AIC of 5 possible models:
1. lm(metabolite~sex)
2. lm(metabolite~genotype)
3. lm(metabolite~sex+genotype)
4. lm(metabolite~sex*genotype)
5. lm(metabolite~1)
```{r,calculate AIC across PCs}
# Create an empty results data frame to store the results
AIC_results <- data.frame(COMPOUND = character(0), 
                          AIC_sex = numeric(0),
                          AIC_geno = numeric(0),
                          AIC_null=numeric(0),
                          AIC_sex_geno=numeric(0),
                          AIC_int=numeric(0))

components<-levels(melted_PC_samples$PC)

# Loop through unique PCs
for (component in components) {
  subset_data <- subset(melted_PC_samples, component == PC)

  if (nrow(subset_data) > 0) {
    null.model <- lm(score ~ 1, data = subset_data, weights=n_samples)
    sex.fixed <- lm(score ~ sex, data = subset_data, weights=n_samples)
    G.only <- lm(score ~ genotype, data = subset_data, weights=n_samples)
    S.plus.G <- lm(score ~ sex + genotype, data = subset_data, weights=n_samples)
    SG.int.full <- lm(score ~ sex*genotype, data = subset_data, weights=n_samples)    
    AIC_sex<-AIC(sex.fixed)
    AIC_geno<-AIC(G.only)
    AIC_null<-AIC(null.model)
    AIC_sex_geno<-AIC(S.plus.G)
    AIC_int<-AIC(SG.int.full)

    # Create a temporary data frame for each PC
    temp_df <- data.frame(component = component, 
    AIC_null=AIC_null,
    AIC_sex=AIC_sex,
    AIC_geno=AIC_geno,
    AIC_sex_geno=AIC_sex_geno,
    AIC_int=AIC_int)

    # Append the temporary data frame to the results
    AIC_results <- distinct(rbind(AIC_results, temp_df))
      }
}

AIC_results$AIC_min <- apply(AIC_results[, 2:6], 1, function(row) {
  min(row)}  # Get the index of the most negative number
)
```
Calculating the delta AIC between the lowest AIC and the next parsimonious model
```{r,calculate delta AIC}
AIC_results <- AIC_results %>%
  mutate(delta_null_s = ifelse(AIC_min == AIC_sex, AIC_null - AIC_min, NA_real_),
         delta_null_g = ifelse(AIC_min == AIC_geno, AIC_null - AIC_min, NA_real_),
         delta_sex = ifelse(AIC_min == AIC_sex_geno, AIC_sex - AIC_sex_geno, NA_real_),
         delta_geno = ifelse(AIC_min == AIC_sex_geno, AIC_geno - AIC_sex_geno, NA_real_),
         delta_sex_geno = ifelse(AIC_min == AIC_int, AIC_sex_geno - AIC_int, NA_real_))
```
Determining the best model for each PC
```{r,determine best model for each PC}
AIC_results_PC <- AIC_results %>%
  mutate(best_fit = case_when(
    AIC_min == AIC_null ~ "AIC_null",
    delta_null_s >= 2 ~ "AIC_sex",
    delta_null_s < 2 ~ "AIC_null",
    delta_null_g >= 2 ~ "AIC_geno",
    delta_null_g < 2 ~ "AIC_null",
    delta_sex >= 2 ~ "AIC_sex_geno",
    delta_sex < 2 ~ "AIC_sex",
    delta_geno >= 2 ~ "AIC_sex_geno",
    delta_geno < 2 ~ "AIC_geno",
    delta_sex_geno >= 2 ~ "AIC_int",
    delta_sex_geno < 2 ~ "AIC_sex_geno",
    TRUE ~ NA_character_
  ))
table(AIC_results_PC$best_fit)
```
Save out results:
```{r}
write.csv(AIC_results_PC,"output/02a_AIC_analysis_by_PC.csv",row.names = F)
```
Saving out .rds files
```{r,save out r files}
save(file='../data/results/AIC_results_PC.data', AIC_results_PC)
rm(list = ls())
```