---
title:   "Sex Variation in the DGRP Metabolome - Univariate"
output:   html_notebook
by:       R. Coig
updated:  5/7/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Univariate analysis, sex/genotype regressions without phenotype
```{r,load required libraries}
library(tidyr)#Data manipulation
library(dplyr)#Data manipulation
library(car)#For specificying Type III ANOVA
```
Loading in metabolome for analysis
```{r,loading in metabolome data}
rm(list = ls())
load('../data/processed/batch_corrected.data')
read.csv("output/02a_AIC_analysis_by_metabolite.csv")->AIC
```
Calculating number of samples per genotype for weighting regressions
```{r}
n_samples <- key %>%
  dplyr::group_by(genotype) %>%
  dplyr::summarise(n_samples = dplyr::n())

melted_dat<-merge(melted_dat,n_samples,by="genotype")
```
Running the sex x genotype Type III ANOVA for all metabolites (Individual Sex)
```{r}
# Create an empty results data frame to store the results
mz_sex_by_geno_results <- data.frame(COMPOUND = character(0), 
                          R_squared = numeric(0),
                          R_adj = numeric(0),
                          sex_p = numeric(0),
                          geno_p=numeric(0),
                          int_p=numeric(0),
                          sex_eta=numeric(0),
                          geno_eta=numeric(0),
                          int_eta=numeric(0))

# Loop through unique mzs
for (metabolite in mz) {
  subset_data <- subset(melted_dat, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
SG.int.full <- lm(value ~ sex*genotype, data = subset_data, weights=n_samples)    
anova_results <- Anova(SG.int.full, type="III")
summary <- summary(SG.int.full)
R_squared<- summary$r.squared
R_adj<-summary$adj.r.squared
sex_p<-anova_results$`Pr(>F)`[2]
geno_p<-anova_results$`Pr(>F)`[3]
int_p<-anova_results$`Pr(>F)`[4]
# Extract Sum Sq for genotype and Residuals
SS_sex <- anova_results$`Sum Sq`[2]
SS_geno <- anova_results$`Sum Sq`[3]
SS_int <- anova_results$`Sum Sq`[4]
SS_total <- sum(anova_results$`Sum Sq`)
# Calculate Eta Squared for genotype
eta_squared_sex <- SS_sex / SS_total
eta_squared_geno <- SS_geno / SS_total
eta_squared_int <- SS_int / SS_total

    # Create a temporary data frame for each metabolite
    temp_df <- data.frame(COMPOUND = metabolite, 
    R_squared=R_squared,
    R_adj=R_adj,
    sex_p=sex_p,
    geno_p=geno_p,
    int_p=int_p,
    sex_eta=eta_squared_sex,
    geno_eta=eta_squared_geno,
    int_eta=eta_squared_int)

    # Append the temporary data frame to the results
    mz_sex_by_geno_results <- distinct(rbind(mz_sex_by_geno_results, temp_df))
      }
}

# Adjust p-values for False Discovery Rate
mz_sex_by_geno_results$sex_p_adj <- p.adjust(mz_sex_by_geno_results$sex_p, method = "fdr")
mz_sex_by_geno_results$geno_p_adj <- p.adjust(mz_sex_by_geno_results$geno_p, method = "fdr")
mz_sex_by_geno_results$int_p_adj <- p.adjust(mz_sex_by_geno_results$int_p, method = "fdr")
```
Running the sex + genotype Type III ANOVA for all metabolites (Individual Sex)
```{r}
# Create an empty results data frame to store the results
mz_sex_plus_geno_results <- data.frame(COMPOUND = character(0), 
                          R_squared = numeric(0),
                          R_adj = numeric(0),
                          sex_p = numeric(0),
                          geno_p=numeric(0),
                          int_p=numeric(0),
                          sex_eta=numeric(0),
                          geno_eta=numeric(0),
                          int_eta=numeric(0))

# Loop through unique mzs
for (metabolite in mz) {
  subset_data <- subset(melted_dat, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
SG.int.full <- lm(value ~ sex + genotype, data = subset_data, weights=n_samples)    
anova_results <- Anova(SG.int.full, type="III")
summary <- summary(SG.int.full)
R_squared<- summary$r.squared
R_adj<-summary$adj.r.squared
sex_p<-anova_results$`Pr(>F)`[2]
geno_p<-anova_results$`Pr(>F)`[3]
#int_p<-anova_results$`Pr(>F)`[4]
# Extract Sum Sq for genotype and Residuals
SS_sex <- anova_results$`Sum Sq`[2]
SS_geno <- anova_results$`Sum Sq`[3]
#SS_int <- anova_results$`Sum Sq`[4]
SS_total <- sum(anova_results$`Sum Sq`)
# Calculate Eta Squared for genotype
eta_squared_sex <- SS_sex / SS_total
eta_squared_geno <- SS_geno / SS_total
#eta_squared_int <- SS_int / SS_total

    # Create a temporary data frame for each metabolite
    temp_df <- data.frame(COMPOUND = metabolite, 
    R_squared=R_squared,
    R_adj=R_adj,
    sex_p=sex_p,
    geno_p=geno_p,
    int_p=NA,
    sex_eta=eta_squared_sex,
    geno_eta=eta_squared_geno,
    int_eta=NA)

    # Append the temporary data frame to the results
    mz_sex_plus_geno_results <- distinct(rbind(mz_sex_plus_geno_results, temp_df))
      }
}

# Adjust p-values for False Discovery Rate
mz_sex_plus_geno_results$sex_p_adj <- p.adjust(mz_sex_plus_geno_results$sex_p, method = "fdr")
mz_sex_plus_geno_results$geno_p_adj <- p.adjust(mz_sex_plus_geno_results$geno_p, method = "fdr")
mz_sex_plus_geno_results$int_p_adj <- NA

```
Running the sex only Type III ANOVA for all metabolites (Population Sex)
```{r}
# Create an empty results data frame to store the results
mz_sex_only_results <- data.frame(COMPOUND = character(0), 
                          R_squared = numeric(0),
                          R_adj = numeric(0),
                          sex_p = numeric(0),
                          geno_p=numeric(0),
                          int_p=numeric(0),
                          sex_eta=numeric(0),
                          geno_eta=numeric(0),
                          int_eta=numeric(0))

# Loop through unique mzs
for (metabolite in mz) {
  subset_data <- subset(melted_dat, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
SG.int.full <- lm(value ~ sex, data = subset_data, weights=n_samples)    
anova_results <- Anova(SG.int.full, type="III")
summary <- summary(SG.int.full)
R_squared<- summary$r.squared
R_adj<-summary$adj.r.squared
sex_p<-anova_results$`Pr(>F)`[2]
#geno_p<-anova_results$`Pr(>F)`[3]
#int_p<-anova_results$`Pr(>F)`[4]
# Extract Sum Sq for genotype and Residuals
SS_sex <- anova_results$`Sum Sq`[2]
#SS_geno <- anova_results$`Sum Sq`[3]
#SS_int <- anova_results$`Sum Sq`[4]
SS_total <- sum(anova_results$`Sum Sq`)
# Calculate Eta Squared for genotype
eta_squared_sex <- SS_sex / SS_total
#eta_squared_geno <- SS_geno / SS_total
#eta_squared_int <- SS_int / SS_total

    # Create a temporary data frame for each metabolite
    temp_df <- data.frame(COMPOUND = metabolite, 
    R_squared=R_squared,
    R_adj=R_adj,
    sex_p=sex_p,
    geno_p=NA,
    int_p=NA,
    sex_eta=eta_squared_sex,
    geno_eta=NA,
    int_eta=NA)

    # Append the temporary data frame to the results
    mz_sex_only_results <- distinct(rbind(mz_sex_only_results, temp_df))
      }
}

# Adjust p-values for False Discovery Rate
mz_sex_only_results$sex_p_adj <- p.adjust(mz_sex_only_results$sex_p, method = "fdr")
mz_sex_only_results$geno_p_adj <- NA
mz_sex_only_results$int_p_adj <- NA
```
Running the genotype only Type III ANOVA for all metabolites 
```{r}
# Create an empty results data frame to store the results
mz_geno_only_results <- data.frame(COMPOUND = character(0), 
                          R_squared = numeric(0),
                          R_adj = numeric(0),
                          sex_p = numeric(0),
                          geno_p=numeric(0),
                          int_p=numeric(0),
                          sex_eta=numeric(0),
                          geno_eta=numeric(0),
                          int_eta=numeric(0))

# Loop through unique mzs
for (metabolite in mz) {
  subset_data <- subset(melted_dat, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
SG.int.full <- lm(value ~ genotype, data = subset_data, weights=n_samples)    
anova_results <- Anova(SG.int.full, type="III")
summary <- summary(SG.int.full)
R_squared<- summary$r.squared
R_adj<-summary$adj.r.squared
#sex_p<-anova_results$`Pr(>F)`[2]
geno_p<-anova_results$`Pr(>F)`[2]
#int_p<-anova_results$`Pr(>F)`[4]
# Extract Sum Sq for genotype and Residuals
#SS_sex <- anova_results$`Sum Sq`[2]
SS_geno <- anova_results$`Sum Sq`[2]
#SS_int <- anova_results$`Sum Sq`[4]
SS_total <- sum(anova_results$`Sum Sq`)
# Calculate Eta Squared for genotype
#eta_squared_sex <- SS_sex / SS_total
eta_squared_geno <- SS_geno / SS_total
#eta_squared_int <- SS_int / SS_total

    # Create a temporary data frame for each metabolite
    temp_df <- data.frame(COMPOUND = metabolite, 
    R_squared=R_squared,
    R_adj=R_adj,
    sex_p=NA,
    geno_p=geno_p,
    int_p=NA,
    sex_eta=NA,
    geno_eta=eta_squared_geno,
    int_eta=NA)

    # Append the temporary data frame to the results
    mz_geno_only_results <- distinct(rbind(mz_geno_only_results, temp_df))
      }
}

# Adjust p-values for False Discovery Rate
mz_geno_only_results$sex_p_adj <- NA
mz_geno_only_results$geno_p_adj <- p.adjust(mz_geno_only_results$geno_p, method = "fdr")
mz_geno_only_results$int_p_adj <- NA
```
Combining results into one df, keeping only the correct results per metabolite by previously determined model.
```{r}
mz_sex_by_geno_results <- merge(AIC[, c(1, 13)], mz_sex_by_geno_results, by = "COMPOUND")
mz_sex_plus_geno_results <- merge(AIC[, c(1, 13)], mz_sex_plus_geno_results, by = "COMPOUND")
mz_sex_only_results <- merge(AIC[, c(1, 13)], mz_sex_only_results, by = "COMPOUND")
mz_geno_only_results <- merge(AIC[, c(1, 13)], mz_geno_only_results, by = "COMPOUND")

# Merge the data frames based on the values in the AIC column
mz_sex_geno_combined_results <- bind_rows(
  mz_sex_by_geno_results %>% filter(best_fit == "AIC_int" | best_fit == "AIC_null"),
  mz_sex_plus_geno_results %>% filter(best_fit == "AIC_sex_geno"),
  mz_geno_only_results %>% filter(best_fit == "AIC_geno"),
  mz_sex_only_results %>% filter(best_fit == "AIC_sex")
)%>% mutate(COMPOUND = factor(COMPOUND, levels = mz)) %>%
  arrange(COMPOUND) %>%
  mutate(COMPOUND = as.character(COMPOUND)) 

mz_sex_geno_combined_results <- mz_sex_geno_combined_results %>%
  mutate(across(-c(COMPOUND, best_fit), ~ if_else(best_fit == "AIC_null", NA, .)))
```
Saving out Univariate results for summary document
```{r}
write.csv(mz_sex_by_geno_results,"output/02b_mz_sex_by_geno_results.csv",row.names=T)
write.csv(mz_sex_plus_geno_results,"output/02b_mz_sex_plus_geno_results.csv",row.names=T)
write.csv(mz_sex_only_results,"output/02b_mz_sex_only_results.csv",row.names=T)
write.csv(mz_geno_only_results,"output/02b_mz_geno_only_results.csv",row.names=T)

write.csv(mz_sex_geno_combined_results,"output/02b_mz_combined_results.csv",row.names=T)
```
Saving out r files
```{r}
save(file='../data/results/mz_sex_geno_results.data', mz_sex_geno_combined_results, mz_sex_only_results, mz)
save(file='../data/results/mz_sex_by_geno_results.data', mz_sex_by_geno_results, mz)
rm(list = ls())
```