---
title:   "Sex Variation in the DGRP Metabolome - PCA"
output:   html_notebook
by:       R. Coig
updated:  5/7/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: PC regression results w/o phenotypes
```{r,load required libraries}
library(tidyr)#Data manipulation
library(dplyr)#Data manipulation
library(car)#For specificying Type III ANOVA
```
Loading in batch corrected data for analysis
```{r,loading in batch corrected metabolome data}
rm(list = ls())
load('../data/processed/PCA.data')
read.csv("output/02a_AIC_analysis_by_PC.csv")->AIC
```
Running the sex x genotype lm regressions for ~ PC 1:25
```{r}
# Create an empty results data frame to store the results
PC_sex_by_geno_results <- data.frame(PC = character(0), 
                          R_squared = numeric(0),
                          R_adj = numeric(0),
                          sex_p = numeric(0),
                          geno_p=numeric(0),
                          int_p=numeric(0),
                          sex_eta=numeric(0),
                          geno_eta=numeric(0),
                          int_eta=numeric(0))

components<-levels(melted_PC_samples$PC)

# Loop through unique PCs
for (component in components) {
  subset_data <- subset(melted_PC_samples, PC == component)

  if (nrow(subset_data) > 0) {
SG.int.full <- lm(score ~ sex*genotype, data = subset_data, weights=n_samples)    
anova_results <- Anova(SG.int.full, type="III")
summary <- summary(SG.int.full)
R_squared<- summary$r.squared
R_adj<-summary$adj.r.squared
sex_p<-anova_results$`Pr(>F)`[2]
geno_p<-anova_results$`Pr(>F)`[3]
int_p<-anova_results$`Pr(>F)`[4]
# Extract Sum Sq for genotype and Residuals
SS_sex <- anova_results$`Sum Sq`[2]
SS_geno <- anova_results$`Sum Sq`[3]
SS_int <- anova_results$`Sum Sq`[4]
SS_total <- sum(anova_results$`Sum Sq`)
# Calculate Eta Squared for genotype
eta_squared_sex <- SS_sex / SS_total
eta_squared_geno <- SS_geno / SS_total
eta_squared_int <- SS_int / SS_total

    # Create a temporary data frame for each PC
    temp_df <- data.frame(PC = component, 
    R_squared=R_squared,
    R_adj=R_adj,
    sex_p=sex_p,
    geno_p=geno_p,
    int_p=int_p,
    sex_eta=eta_squared_sex,
    geno_eta=eta_squared_geno,
    int_eta=eta_squared_int)

    # Append the temporary data frame to the results
    PC_sex_by_geno_results <- distinct(rbind(PC_sex_by_geno_results, temp_df))
      }
}

# Adjust p-values for False Discovery Rate
PC_sex_by_geno_results$sex_p_adj <- p.adjust(PC_sex_by_geno_results$sex_p, method = "fdr")
PC_sex_by_geno_results$geno_p_adj <- p.adjust(PC_sex_by_geno_results$geno_p, method = "fdr")
PC_sex_by_geno_results$int_p_adj <- p.adjust(PC_sex_by_geno_results$int_p, method = "fdr")
```
Running the sex + genotype lm regressions for ~ PC 1:25
```{r}
# Create an empty results data frame to store the results
PC_sex_plus_geno_results <- data.frame(PC = character(0), 
                          R_squared = numeric(0),
                          R_adj = numeric(0),
                          sex_p = numeric(0),
                          geno_p=numeric(0),
                          sex_eta=numeric(0),
                          geno_eta=numeric(0))

components<-levels(melted_PC_samples$PC)

# Loop through unique PCs
for (component in components) {
  subset_data <- subset(melted_PC_samples, PC == component)

  if (nrow(subset_data) > 0) {
SG.int.full <- lm(score ~ sex + genotype, data = subset_data, weights=n_samples)    
anova_results <- Anova(SG.int.full, type="III")
summary <- summary(SG.int.full)
R_squared<- summary$r.squared
R_adj<-summary$adj.r.squared
sex_p<-anova_results$`Pr(>F)`[2]
geno_p<-anova_results$`Pr(>F)`[3]
#int_p<-anova_results$`Pr(>F)`[4]
# Extract Sum Sq for genotype and Residuals
SS_sex <- anova_results$`Sum Sq`[2]
SS_geno <- anova_results$`Sum Sq`[3]
#SS_int <- anova_results$`Sum Sq`[4]
SS_total <- sum(anova_results$`Sum Sq`)
# Calculate Eta Squared for genotype
eta_squared_sex <- SS_sex / SS_total
eta_squared_geno <- SS_geno / SS_total
#eta_squared_int <- SS_int / SS_total

    # Create a temporary data frame for each PC
    temp_df <- data.frame(PC = component, 
    R_squared=R_squared,
    R_adj=R_adj,
    sex_p=sex_p,
    geno_p=geno_p,
    int_p=NA,
    sex_eta=eta_squared_sex,
    geno_eta=eta_squared_geno,
    int_eta=NA)

    # Append the temporary data frame to the results
    PC_sex_plus_geno_results <- distinct(rbind(PC_sex_plus_geno_results, temp_df))
      }
}

# Adjust p-values for False Discovery Rate
PC_sex_plus_geno_results$sex_p_adj <- p.adjust(PC_sex_plus_geno_results$sex_p, method = "fdr")
PC_sex_plus_geno_results$geno_p_adj <- p.adjust(PC_sex_plus_geno_results$geno_p, method = "fdr")
PC_sex_plus_geno_results$int_p_adj <- NA

```
Running the sex only lm regressions for ~ PC 1:25
```{r}
# Create an empty results data frame to store the results
PC_sex_only_results <- data.frame(PC = character(0), 
                          R_squared = numeric(0),
                          R_adj = numeric(0),
                          sex_p = numeric(0),
                          geno_p=numeric(0),
                          sex_eta=numeric(0),
                          geno_eta=numeric(0))

components<-levels(melted_PC_samples$PC)

# Loop through unique PCs
for (component in components) {
  subset_data <- subset(melted_PC_samples, PC == component)

  if (nrow(subset_data) > 0) {
SG.int.full <- lm(score ~ sex, data = subset_data, weights=n_samples)    
anova_results <- Anova(SG.int.full, type="III")
summary <- summary(SG.int.full)
R_squared<- summary$r.squared
R_adj<-summary$adj.r.squared
sex_p<-anova_results$`Pr(>F)`[2]
#geno_p<-anova_results$`Pr(>F)`[3]
#int_p<-anova_results$`Pr(>F)`[4]
# Extract Sum Sq for genotype and Residuals
SS_sex <- anova_results$`Sum Sq`[2]
#SS_geno <- anova_results$`Sum Sq`[3]
#SS_int <- anova_results$`Sum Sq`[4]
SS_total <- sum(anova_results$`Sum Sq`)
# Calculate Eta Squared for genotype
eta_squared_sex <- SS_sex / SS_total
#eta_squared_geno <- SS_geno / SS_total
#eta_squared_int <- SS_int / SS_total

    # Create a temporary data frame for each PC
    temp_df <- data.frame(PC = component, 
    R_squared=R_squared,
    R_adj=R_adj,
    sex_p=sex_p,
    geno_p=NA,
    int_p=NA,
    sex_eta=eta_squared_sex,
    geno_eta=NA,
    int_eta=NA)

    # Append the temporary data frame to the results
    PC_sex_only_results <- distinct(rbind(PC_sex_only_results, temp_df))
      }
}

# Adjust p-values for False Discovery Rate
PC_sex_only_results$sex_p_adj <- p.adjust(PC_sex_only_results$sex_p, method = "fdr")
PC_sex_only_results$geno_p_adj <- NA
PC_sex_only_results$int_p_adj <- NA
```
Running the genotype only lm regressions for ~ PC 1:25
```{r}
# Create an empty results data frame to store the results
PC_geno_only_results <- data.frame(PC = character(0), 
                          R_squared = numeric(0),
                          R_adj = numeric(0),
                          sex_p = numeric(0),
                          geno_p=numeric(0),
                          sex_eta=numeric(0),
                          geno_eta=numeric(0))

components<-levels(melted_PC_samples$PC)

# Loop through unique PCs
for (component in components) {
  subset_data <- subset(melted_PC_samples, PC == component)

  if (nrow(subset_data) > 0) {
SG.int.full <- lm(score ~ genotype, data = subset_data, weights=n_samples)    
anova_results <- Anova(SG.int.full, type="III")
summary <- summary(SG.int.full)
R_squared<- summary$r.squared
R_adj<-summary$adj.r.squared
#sex_p<-anova_results$`Pr(>F)`[2]
geno_p<-anova_results$`Pr(>F)`[2]
#int_p<-anova_results$`Pr(>F)`[4]
# Extract Sum Sq for genotype and Residuals
#SS_sex <- anova_results$`Sum Sq`[2]
SS_geno <- anova_results$`Sum Sq`[2]
#SS_int <- anova_results$`Sum Sq`[4]
SS_total <- sum(anova_results$`Sum Sq`)
# Calculate Eta Squared for genotype
#eta_squared_sex <- SS_sex / SS_total
eta_squared_geno <- SS_geno / SS_total
#eta_squared_int <- SS_int / SS_total

    # Create a temporary data frame for each PC
    temp_df <- data.frame(PC = component, 
    R_squared=R_squared,
    R_adj=R_adj,
    sex_p=NA,
    geno_p=geno_p,
    int_p=NA,
    sex_eta=NA,
    geno_eta=eta_squared_geno,
    int_eta=NA)

    # Append the temporary data frame to the results
    PC_geno_only_results <- distinct(rbind(PC_geno_only_results, temp_df))
      }
}

# Adjust p-values for False Discovery Rate
PC_geno_only_results$sex_p_adj <- NA
PC_geno_only_results$geno_p_adj <- p.adjust(PC_geno_only_results$geno_p, method = "fdr")
PC_geno_only_results$int_p_adj <- NA
```
Combining results into one df, keeping only the correct results per PC by previously determined model.
```{r}
PC_sex_by_geno_results <- merge(AIC[, c(1, 13)], PC_sex_by_geno_results, by.x = "component", by.y = "PC")
PC_sex_plus_geno_results <- merge(AIC[, c(1, 13)], PC_sex_plus_geno_results, by.x = "component", by.y = "PC")
PC_sex_only_results <- merge(AIC[, c(1, 13)], PC_sex_only_results, by.x = "component", by.y = "PC")
PC_geno_only_results <- merge(AIC[, c(1, 13)], PC_geno_only_results, by.x = "component", by.y = "PC")

# Merge the data frames based on the values in the AIC column
PC_sex_geno_combined_results <- bind_rows(
  PC_sex_by_geno_results %>% filter(best_fit == "AIC_int"),
  PC_sex_plus_geno_results %>% filter(best_fit == "AIC_sex_geno"),
  PC_geno_only_results %>% filter(best_fit == "AIC_geno")
)%>% mutate(component = factor(component, levels = components)) %>%
  arrange(component)
```
Saving out PCA results for summary document
```{r}
write.csv(PC_sex_by_geno_results,"output/02b_PC_sex_by_geno_results.csv",row.names=T)
write.csv(PC_sex_plus_geno_results,"output/02b_PC_sex_plus_geno_results.csv",row.names=T)
write.csv(PC_sex_only_results,"output/02b_PC_sex_only_results.csv",row.names=T)
write.csv(PC_geno_only_results,"output/02b_PC_geno_only_results.csv",row.names=T)

write.csv(PC_sex_geno_combined_results,"output/02b_PC_combined_results.csv",row.names=T)
```
Saving out r files
```{r}
save(file='../data/results/PCA_sex_geno_results.data', PC_sex_geno_combined_results, PC_sex_only_results, components)
rm(list = ls())
```