---
title:   "Sex Variation in the DGRP Metabolome - Figures"
output:   html_notebook
by:       R. Coig
updated:  3/31/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Figure 1-Mass
```{r,load required libraries}
library(dplyr)
library(ggplot2)
```
Loading summary statistics and calculated dimorphism metrics for TAG and mass
```{r,loading in phenotype data}
rm(list = ls())
load('../data/processed/phenotype-mass.data')
load('../data/processed/phenotype-dimorphism.data')
dimorphism_summary<-merge(mass_sex_geno_wide[,c(1,21)],dimorphism_summary,by="genotype")%>%filter(!is.na(diff_mass))
```
Calculate means of SD phenotypes
```{r}
mean_mass_SD4<-round(mean(dimorphism_summary$diff_mass,na.rm=T),2)
```
mass DGRP plot 
```{r}
dimorphism_summary<-dimorphism_summary%>%dplyr::arrange(desc(diff_mass))%>%
  mutate(genotype = factor(genotype, levels = unique(genotype)))
dimorphism_summary <- dimorphism_summary %>%
  mutate(genotype_order = as.numeric(factor(genotype)))

genos_right <- dimorphism_summary %>%
  filter(diff_mass > mean_mass_SD4) %>% pull(genotype)%>%droplevels()

genos_missing_se<- dimorphism_summary %>%
  filter(is.na(mass_4_pl_se)) %>% pull(genotype)%>%droplevels()


ggplot(dimorphism_summary, aes(x = genotype, y = diff_mass)) +
  geom_hline(yintercept = mean_mass_SD4, color = "lightgray", linetype = "dashed") +
    geom_errorbar(aes(ymin = diff_mass - mass_4_pl_se, ymax = diff_mass + mass_4_pl_se), width = 0.2) +  # Add error bars
  geom_point(aes(fill = diff_mass), size = 3.5, stroke = 1, shape = 21) +
  labs(x = "DGRP Genotype", y = "\u0394Mass") +
geom_text(data = dimorphism_summary %>% filter(diff_mass > mean_mass_SD4),aes(label = genotype,y=diff_mass+mass_4_pl_se+.03), size = 4, check_overlap=F)+
geom_text(data = dimorphism_summary %>% filter(diff_mass < mean_mass_SD4),aes(label = genotype, y=diff_mass-mass_4_pl_se-.03), size = 4, check_overlap = FALSE) + 
  geom_text(data = dimorphism_summary %>% filter(genotype%in%genos_missing_se),aes(label = genotype, y=diff_mass+.05), size = 4, check_overlap = FALSE) + 
  scale_fill_gradient(
    low = "#2B9DEE", high = "#F7DA6E",
    guide = FALSE  # Remove the existing fill legend
  ) +
  ylim(0.4,1.7)+
  theme_classic() +
  theme(
    axis.title = element_text(size = 20),
    axis.text.x = element_blank(),
    axis.text = element_text(size = 16),
    axis.ticks.x = element_blank(),
    legend.position = "none"  # Remove the default legend
  )
#+scale_y_continuous(labels = function(x) sprintf("%.4f", x))


ggsave("plots/Fig1B-mass.jpg", width = 9, height = 6)
```

Female Mass mu by Male Mass mu
```{r}
# Calculate Spearman's correlation coefficient and p-value
spearman_cor_mass4 <- cor.test(dimorphism_summary$mass_4_mu_ss_male, dimorphism_summary$mass_4_mu_ss_female, method = "spearman")
cor_coef_mass4 <- round(spearman_cor_mass4$estimate,2)
p_value_mass4 <- round(spearman_cor_mass4$p.value,11)

# Plot with annotations
# Plot with annotations
ggplot(dimorphism_summary, aes(x = mass_4_mu_ss_male, y = mass_4_mu_ss_female, label = genotype, fill = diff_mass)) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "solid") +  # Add the line of unity 
  geom_segment(data = subset(dimorphism_summary, genotype == 535),
               aes(x = mass_4_mu_ss_male, y=mass_4_mu_ss_male, xend = mass_4_mu_ss_male, yend = mass_4_mu_ss_female), 
               color = "black", size = 1) +
  geom_point(size = 8, pch = 21, stroke = 1) +
  geom_text(data = subset(dimorphism_summary, genotype == 535| genotype == 31),size = 5, nudge_y=.1) +
  labs(
    x = "Mean male mass",
    y = "Mean female mass"
  ) +
  scale_fill_gradient(high = "#F7DA6E", low = "#2B9DEE", guide = "none") +
  theme_classic() +
  coord_fixed() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title=element_text(size=20),
        axis.text=element_text(size=16)) +
  geom_text(x = 2.4,
            y = 5,
            label = paste("ρ =", cor_coef_mass4, "\np =", p_value_mass4),
            size = 8, hjust = 0, vjust = 1, color = "black")+
    geom_text(data = subset(dimorphism_summary, genotype == 535),
            aes(x = 3.2, y = 3.9, 
                label = "\u0394Mass"),
            size = 8, hjust = 0, vjust = 0, color = "black")#+
#geom_segment(
 #   aes(x = 3.6, y = 3.51, xend = 3.55, yend = 3.51),
  #  arrow = arrow(length = unit(0.1, "inches")), # Adjust the length of the arrow as needed
   # color = "black"
#  )

# Save the plot to a PNG file with specific dimensions
ggsave("plots/Fig1A-mass.jpg", width = 5, height = 7)
```