---
title:   "Sex Variation in the DGRP Metabolome - Preprocess"
output:   html_notebook
by:       R. Coig
updated:  5/7/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Normalizing the raw metabolome data.
```{r,load required libraries}
library(reshape2)
library(dplyr)
library(tidyr)
library(pdftools)
library(purrr)
library(Rmisc)
library(ggplot2)
```
Loading in raw data. The raw data frame has data for 198 metabolites, including QC samples. QC(S) samples have all the same compounds as the experimental samples, and the same sample was run on the instrument after every 10 experimental samples. 
```{r,loading in mz and QC data}
rm(list = ls())
load('data/raw/raw.data')
```
We start by creating log transformed and scaled dataframes
```{r,log transforming the data}
# Log-transform the data
log_mz <- log10(raw)
log_by_sample<-as.data.frame(t(log_mz))

boxplot(log10((raw)))
boxplot(t(log_by_sample))
```
Adding keys for plotting/analysis.
```{r,}
log_key<-log_by_sample
log_key$order<-row.names(log_key)
log_key<-merge(key,log_key,by="order")
log_key <- log_key %>% arrange(run_order)

melted_log <- melt(log_key[,c(3,14:198)],id.vars = "run_order")
colnames(melted_log) <- c("run_order", "COMPOUND", "value")
melted_log <- merge(melted_log,key,by="run_order")

ggplot(melted_log,aes(x=order,y=value, color=sex))+
geom_point()+
  geom_smooth(aes(color=sex),method="lm")+
  facet_grid(~assayer+sex)

ggplot(melted_raw,aes(x=order,y=value, color=sex))+
geom_point()+
  geom_smooth(aes(color=sex),method="lm")+
  facet_grid(~assayer+sex)

ggplot(melted_log,aes(x=order,y=value, color=sex))+
geom_point()+
  geom_smooth(aes(color=sex),method="lm")+
  facet_grid(~sex+box+assayer)

ggplot(melted_log,aes(x=run_order,y=value, color=sex))+
geom_point()+
  geom_smooth(aes(color=sex),method="lm")
```
We find that LH's male samples are lower because of assayer protocol differences in tissue collection, before scaling we will median center these four groups in the log corrected data
```{r}
# Calculate median for each group using dplyr
medians <- melted_log %>%
  group_by(assayer,sex) %>%
  dplyr::summarize(assayer=assayer,
            sex=sex,
            median_value = median(value))%>%distinct()
medians$median_value<-medians$median_value-min(medians$median_value)

# Merge median values back to the original data
melted_adjusted <- melted_log %>%
  left_join(medians, by = c("assayer", "sex")) %>%
  mutate(value = value - median_value)

ggplot(melted_adjusted,aes(x=order,y=value, color=sex))+
geom_point()+
  geom_smooth(aes(color=sex),method="lm")+
  facet_grid(~assayer+sex+box)

ggplot(melted_adjusted,aes(x=run_order,y=value, color=sex))+
geom_point()+
  geom_smooth(aes(color=sex),method="lm")
```
Now the medians of all groups are the same.
```{r}
melted_adjusted<-melted_adjusted[,1:3]
adjusted_key <- dcast(melted_adjusted, run_order ~ COMPOUND, value.var = "value")
adjusted_by_sample<-adjusted_key[,-1]
```


```{r,}
scale_by_sample <- as.data.frame(t(apply(adjusted_by_sample, 1, function(x) scale(x, center = TRUE, scale = TRUE))))
colnames(scale_by_sample)<-mz

scale_key<-scale_by_sample
scale_key$run_order<-row.names(scale_key)
scale_key<-merge(key,scale_key,by="run_order")
scale_key <- scale_key %>% arrange(run_order)

melted_scale <- melt(scale_key[,c(1,14:198)],id.vars = "run_order")
colnames(melted_scale) <- c("run_order", "COMPOUND", "value")
melted_scale <- merge(melted_scale,key,by="run_order")

ggplot(melted_scale,aes(x=order,y=value, color=sex))+
geom_point()+
  geom_smooth(aes(color=sex),method="lm")+
  facet_grid(~assayer+sex)

ggplot(melted_scale,aes(x=run_order,y=value, color=sex))+
geom_point()+
  geom_smooth(aes(color=sex),method="lm")
```
Plotting metabolites across run order (log10)
```{r}
# Create a list of ggplot objects for each COMPOUND
plots_list <- melted_log %>%
  group_split(COMPOUND) %>%
  map(function(subset) {
    ggplot(subset, aes(x = run_order, y = value, color = batch,shape=type)) +
      geom_point() + 
      geom_line(data = subset %>% filter(type == "pooled fly heads")) +
      theme_classic() +
      ggtitle(paste("Compound:", unique(subset$COMPOUND)))+
      guides(shape = FALSE)
  })

# Function to save plots to multiple pages in a PDF file
save_plots_to_pdf <- function(plots_list, file_name_prefix, plots_per_page) {
  num_plots <- length(plots_list)
  num_pages <- ceiling(num_plots / plots_per_page)
  
  for (i in 1:num_pages) {
    start_idx <- (i - 1) * plots_per_page + 1
    end_idx <- min(i * plots_per_page, num_plots)
    plots_on_page <- plots_list[start_idx:end_idx]
    
    pdf(paste0(file_name_prefix, i, ".pdf"), width = 11, height = 8.5) # Adjust width and height as needed
    
    multiplot(plotlist = plots_on_page, cols = 1) # 1 metabolite per row
    
    dev.off()
  }
}

# Save the plots to multiple pages in a PDF file with 5 plots per page
save_plots_to_pdf(plots_list, "plots_page_", plots_per_page = 5)

# List of file names for the individual PDFs
pdf_files <- paste0("plots_page_", 1:37, ".pdf")

# Output file name for the combined PDF
output_pdf <- "plots/00a_log10_metabolites_by_run_order.pdf"

# Combine the individual PDFs into a single PDF
pdf_combine(pdf_files, output_pdf)

# Clean up: Delete the individual PDF files if needed
file.remove(pdf_files)

```
Plotting metabolites across run order (log and scale, mean-centered)
```{r}
# Create a list of ggplot objects for each COMPOUND
plots_list <- melted_scale %>%
  group_split(COMPOUND) %>%
  map(function(subset) {
    ggplot(subset, aes(x = run_order, y = value, color = batch,shape=type)) +
      geom_point() + 
      geom_line(data = subset %>% filter(type == "pooled fly heads")) +
      theme_classic() +
      ggtitle(paste("Compound:", unique(subset$COMPOUND)))+
      guides(shape = FALSE)
  })

# Function to save plots to multiple pages in a PDF file
save_plots_to_pdf <- function(plots_list, file_name_prefix, plots_per_page) {
  num_plots <- length(plots_list)
  num_pages <- ceiling(num_plots / plots_per_page)
  
  for (i in 1:num_pages) {
    start_idx <- (i - 1) * plots_per_page + 1
    end_idx <- min(i * plots_per_page, num_plots)
    plots_on_page <- plots_list[start_idx:end_idx]
    
    pdf(paste0(file_name_prefix, i, ".pdf"), width = 11, height = 8.5) # Adjust width and height as needed
    
    multiplot(plotlist = plots_on_page, cols = 1) # 1 metabolite per row
    
    dev.off()
  }
}

# Save the plots to multiple pages in a PDF file with 5 plots per page
save_plots_to_pdf(plots_list, "plots_page_", plots_per_page = 5)

# List of file names for the individual PDFs
pdf_files <- paste0("plots_page_", 1:37, ".pdf")

# Output file name for the combined PDF
output_pdf <- "plots/00a_log10_scaled_metabolites_by_run_order.pdf"

# Combine the individual PDFs into a single PDF
pdf_combine(pdf_files, output_pdf)

# Clean up: Delete the individual PDF files if needed
file.remove(pdf_files)

```
Plotting metabolites across run order (log and scale, mean-centered) colored by assayer
```{r}
# Create a list of ggplot objects for each COMPOUND
plots_list <- melted_scale %>%
  group_split(COMPOUND) %>%
  map(function(subset) {
    ggplot(subset, aes(x = run_order, y = value, color = assayer,shape=type)) +
      scale_color_manual(values=c("red","black"))+
      geom_point() + 
      geom_line(data = subset %>% filter(type == "pooled fly heads")) +
      theme_classic() +
      ggtitle(paste("Compound:", unique(subset$COMPOUND)))+
      guides(shape = FALSE)
  })

# Function to save plots to multiple pages in a PDF file
save_plots_to_pdf <- function(plots_list, file_name_prefix, plots_per_page) {
  num_plots <- length(plots_list)
  num_pages <- ceiling(num_plots / plots_per_page)
  
  for (i in 1:num_pages) {
    start_idx <- (i - 1) * plots_per_page + 1
    end_idx <- min(i * plots_per_page, num_plots)
    plots_on_page <- plots_list[start_idx:end_idx]
    
    pdf(paste0(file_name_prefix, i, ".pdf"), width = 11, height = 8.5) # Adjust width and height as needed
    
    multiplot(plotlist = plots_on_page, cols = 1) # 1 metabolite per row
    
    dev.off()
  }
}

# Save the plots to multiple pages in a PDF file with 5 plots per page
save_plots_to_pdf(plots_list, "plots_page_", plots_per_page = 5)

# List of file names for the individual PDFs
pdf_files <- paste0("plots_page_", 1:37, ".pdf")

# Output file name for the combined PDF
output_pdf <- "plots/00a_metabolites_by_run_order_colored_assayer.pdf"

# Combine the individual PDFs into a single PDF
pdf_combine(pdf_files, output_pdf)

# Clean up: Delete the individual PDF files if needed
file.remove(pdf_files)

```
Plotting metabolites across run order (log and scale, mean-centered) colored by sex
```{r}
# Create a list of ggplot objects for each COMPOUND
plots_list <- melted_scale %>%
  group_split(COMPOUND) %>%
  map(function(subset) {
    ggplot(subset, aes(x = run_order, y = value, color = sex,shape=type)) +
      scale_color_manual(values=c("turquoise","purple","gray"))+
      geom_point() + 
      geom_line(data = subset %>% filter(type == "pooled fly heads")) +
      theme_classic() +
      ggtitle(paste("Compound:", unique(subset$COMPOUND)))+
      guides(shape = FALSE)
  })

# Function to save plots to multiple pages in a PDF file
save_plots_to_pdf <- function(plots_list, file_name_prefix, plots_per_page) {
  num_plots <- length(plots_list)
  num_pages <- ceiling(num_plots / plots_per_page)
  
  for (i in 1:num_pages) {
    start_idx <- (i - 1) * plots_per_page + 1
    end_idx <- min(i * plots_per_page, num_plots)
    plots_on_page <- plots_list[start_idx:end_idx]
    
    pdf(paste0(file_name_prefix, i, ".pdf"), width = 11, height = 8.5) # Adjust width and height as needed
    
    multiplot(plotlist = plots_on_page, cols = 1) # 1 metabolite per row
    
    dev.off()
  }
}

# Save the plots to multiple pages in a PDF file with 5 plots per page
save_plots_to_pdf(plots_list, "plots_page_", plots_per_page = 5)

# List of file names for the individual PDFs
pdf_files <- paste0("plots_page_", 1:37, ".pdf")

# Output file name for the combined PDF
output_pdf <- "plots/00a_metabolites_by_run_order_colored_sex.pdf"

# Combine the individual PDFs into a single PDF
pdf_combine(pdf_files, output_pdf)

# Clean up: Delete the individual PDF files if needed
file.remove(pdf_files)

```
Removing QC(S) samples
```{r}
scale_key<-scale_key%>%filter(!is.na(assayer))
scale_key$order<-as.numeric(as.character(scale_key$order))
scale_key$genotype<-droplevels(scale_key$genotype)
scale_key$sex<-droplevels(scale_key$sex)
scale_key$batch<-droplevels(scale_key$batch)
scale_key$type<-NULL
scale_key$run_order<-NULL
scale_key$complete_mzs<-NULL
melted_scale <- melt(scale_key[,c(1,11:195)],id.vars = "order")
colnames(melted_scale) <- c("order", "COMPOUND", "value")
melted_scale <- merge(melted_scale,key,by="order")
scale_by_sample<-scale_key[,-c(1:10)]
key<-scale_key[,c(1:10)]
```
Saving out normalized files
```{r,save out csv files}
write.csv(scale_by_sample,"data/processed/00a_normalized_metabolites.csv",row.names=T)
```
Saving out r files
```{r,save out r files}
save(file='data/processed/normalized.data', scale_by_sample, scale_key, melted_scale, mz, key)
rm(list=ls())
```
