---
title:   "Sex Variation in the DGRP Metabolome - Preprocess"
output:   html_notebook
by:       R. Coig
updated:  3/31/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Batch Correction
```{r,load required libraries}
library(reshape2)
library(dplyr)
library(tidyr)
library(pdftools)
library(purrr)
library(Rmisc)
library(ggplot2)
library(factoextra)#for PCA
```
Loading in log, centered and scaled data
```{r,loading in scaled data}
rm(list = ls())
load('data/processed/normalized.data')
```
PCA plots prior to batch correction to identify batch variables
```{r}
scaled_PCA <- prcomp(scale_by_sample,scale. = T)
sex<-scale_key$sex
batch<-scale_key$batch
box<-scale_key$box
genotype<-scale_key$genotype
assayer<-scale_key$assayer
```
Plotting PCAs of Batch effects in scaled data
```{r}
fviz_pca_ind(scaled_PCA,
             col.ind = factor(batch), # color by groups
             #palette = c("black",  "red"),
             addEllipses = TRUE, # Concentration ellipses
             axes = c(1,2),
             ellipse.type = "confidence",
             legend.title = "",
             repel = TRUE,
             label = "none",
             title = "Batch Effects before correction")+
  theme(plot.title = element_text(hjust = 0.5),legend.position = "bottom",
        panel.background = element_rect(fill = "white"))

fviz_pca_ind(scaled_PCA,
             col.ind = factor(sex), # color by groups
             palette = c("turquoise",  "purple","black"),
             addEllipses = TRUE, # Concentration ellipses
             axes = c(1,2),
             ellipse.type = "confidence",
             legend.title = "",
             repel = TRUE,
             label = "none",
             title = "Sex Effects before correction")+
  theme(plot.title = element_text(hjust = 0.5),legend.position = "bottom",
        panel.background = element_rect(fill = "white"))

fviz_pca_ind(scaled_PCA,
             col.ind = factor(box), # color by groups
             #palette = c("black",  "red"),
             addEllipses = TRUE, # Concentration ellipses
             axes = c(1,2),
             ellipse.type = "confidence",
             legend.title = "",
             repel = TRUE,
             label = "none",
             title = "Box Effects before correction")+
  theme(plot.title = element_text(hjust = 0.5),legend.position = "bottom",
        panel.background = element_rect(fill = "white"))
```
We are expecting a typical effect of Batch (preparation of samples at the LC-MS lab). It looks like there is also an effect of Box (samples were stored in 2 different boxes), but that may be a function of Batch. We will assess the main effects of these variables in the data across metabolites:
Main effect of Batch:90/185(49%) metabolites affected
```{r}
# Create an empty data frame to store the order information
order_info <- data.frame(COMPOUND = character(0), Order = integer(0))
# Create an empty results data frame to store the results
batch_results <- data.frame(COMPOUND = character(0), B_p_value = numeric(0))

# Create an empty residuals data frame to store the correction
batch_correction <- as.data.frame(matrix(nc=185, nr=141))
colnames(batch_correction)<-mz

# Loop through unique metabolite values
for (metabolite in mz) {
  subset_data <- subset(melted_scale, COMPOUND == metabolite)
subset_data<-subset_data%>%arrange(order)
  if (nrow(subset_data) > 0) {
    model <- lm(value ~ batch, data = subset_data)
    B_p_value <- anova(model)$`Pr(>F)`[1]
    
    # Create a temporary data frame for each metabolite
    temp_df <- data.frame(COMPOUND = metabolite, B_p_value = B_p_value)

    # Append the temporary data frame to the results
    batch_results <- rbind(batch_results, temp_df)
    
      # Calculate residuals
    residuals <- model$residuals
    order_info_subset <- data.frame(COMPOUND = metabolite, Order = subset_data$order)
    
    # Append the order information to the order_info data frame
    order_info <- rbind(order_info, order_info_subset)
    
    # Sort residuals based on order
    batch_main_residuals <- residuals[subset_data$order]
    
    # Add residuals to the residuals_df data frame, naming the column with the metabolite name
    batch_correction[,metabolite] <- batch_main_residuals

  }
}

# Sort the results by B_p_value
batch_results <- batch_results %>% arrange(B_p_value)
any(is.na(batch_results))#Check for NAs

# Add adjusted p-values using the BH method
batch_results <- batch_results %>%
  mutate(B_p_value_adjusted = p.adjust(B_p_value, method = "BH"))
sig_batch_main <- batch_results$COMPOUND[batch_results$B_p_value_adjusted < 0.05]
```
Melting the residuals for plotting and analysis
```{r}
BA_key<-batch_correction
BA_key$order<-order_info_subset$Order
BA_key<-merge(key,BA_key,by="order")
melted_BA <- melt(BA_key[,c(1,11:195)],id.vars = "order")
colnames(melted_BA) <- c("order", "COMPOUND", "value")
melted_BA <- merge(melted_BA,key,by="order")
```
Plotting PCAs of Batch effects in corrected data
```{r}
corrected_PCA <- prcomp(batch_correction,scale. = T)

fviz_pca_ind(corrected_PCA,
             col.ind = factor(batch), # color by groups
             #palette = c("black",  "red"),
             addEllipses = TRUE, # Concentration ellipses
             axes = c(1,2),
             ellipse.type = "confidence",
             legend.title = "",
             repel = TRUE,
             label = "none",
             title = "Batch Effects after correction")+
  theme(plot.title = element_text(hjust = 0.5),legend.position = "bottom",
        panel.background = element_rect(fill = "white"))

fviz_pca_ind(corrected_PCA,
             col.ind = factor(sex), # color by groups
             palette = c("turquoise",  "purple","black"),
             addEllipses = TRUE, # Concentration ellipses
             axes = c(1,2),
             ellipse.type = "confidence",
             legend.title = "",
             repel = TRUE,
             label = "none",
             title = "Sex Effects after correction")+
  theme(plot.title = element_text(hjust = 0.5),legend.position = "bottom",
        panel.background = element_rect(fill = "white"))

fviz_pca_ind(corrected_PCA,
             col.ind = factor(box), # color by groups
             #palette = c("black",  "red"),
             addEllipses = TRUE, # Concentration ellipses
             axes = c(1,2),
             ellipse.type = "confidence",
             legend.title = "",
             repel = TRUE,
             label = "none",
             title = "Box Effects after correction")+
  theme(plot.title = element_text(hjust = 0.5),legend.position = "bottom",
        panel.background = element_rect(fill = "white"))
# Specify the file name and dimensions
file_name <- "plots/00b_PCA_corrected_box.jpg"
width <- 8  # Width in inches
height <- 6  # Height in inches
```
Plotting Assayer effects after initial batch correction
```{r}
ggplot(melted_BA,aes(x=order,y=value, color=sex))+
geom_point()+
  geom_smooth(aes(color=sex))+
  facet_grid(~assayer+sex)

ggplot(melted_scale,aes(x=order,y=value, color=sex))+
geom_point()+
  geom_smooth(aes(color=sex))+
  facet_grid(~assayer+sex)

ggplot(melted_scale,aes(x=order,y=value, color=sex))+
geom_point()+
  geom_smooth(aes(color=sex))+
  facet_grid(~box)

ggplot(melted_scale,aes(x=order,y=value))+
geom_point()+
  geom_smooth(aes(color=batch))
```
Preparing final batch corrected dfs for export
```{r}
dat_key<-BA_key

dat_by_sample<-dat_key[,11:195]
key<-dat_key[,1:10]
melted_dat <- melt(dat_key[,c(1,11:195)],id.vars = "order")
colnames(melted_dat) <- c("order", "COMPOUND", "value")
melted_dat <- merge(melted_dat,key,by="order")
```
Saving out batch corrected files for analysis.
```{r,save out batch corrected files}
write.csv(dat_by_sample,"data/processed/corrected_metabolites.csv",row.names=T)

save(file='../data/processed/batch_corrected.data', dat_by_sample, dat_key, melted_dat, mz,  key)
rm(list=ls())
```