---
title:   "Sex Variation in the DGRP Metabolome - Univariate analysis"
output:   html_notebook
by:       R. Coig
updated:  5/7/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Univariate analysis, Within Sex TAG-Metabolite level correlations (All Samples)
```{r,load required libraries}
library(tidyr)#Data manipulation
library(dplyr)#Data manipulation
library(smatr)#Computing relative TAG
library(reshape2)
```
Loading in batch corrected data for analysis
```{r,loading in batch corrected metabolome data}
rm(list = ls())
load('../data/processed/batch_corrected.data')
load('../data/processed/mass_for_correlations.data')
```
TAG Correlations across all metabolites for females
```{r}
F_dat<-melted_dat%>%filter(sex=="female")

# Create an empty results data frame to store the results
FTAG_mz_COR_results <- data.frame(COMPOUND = character(0), 
                                    FTAG_R_spearman = numeric(0),
                                    FTAG_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(F_dat, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$absn5, subset_data$value, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          FTAG_R_spearman = spearman_result$estimate,
                          FTAG_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    FTAG_mz_COR_results <- distinct(rbind(FTAG_mz_COR_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
FTAG_mz_COR_results$FTAG_adj <- p.adjust(FTAG_mz_COR_results$FTAG_p, method = "fdr")
FTAG_mz_COR_results<-FTAG_mz_COR_results%>%arrange(FTAG_p)
FTAG_mz_COR_results$sig_FTAG <- ifelse(FTAG_mz_COR_results$FTAG_adj < 0.01, "SIG", "NS")

sig_FTAG_mz<-FTAG_mz_COR_results%>%filter(sig_FTAG=="SIG")%>%pull(COMPOUND)
```
TAG Correlations across all metabolites for males
```{r}
M_dat<-melted_dat%>%filter(sex=="male")

# Create an empty results data frame to store the results
MTAG_mz_COR_results <- data.frame(COMPOUND = character(0), 
                                    MTAG_R_spearman = numeric(0),
                                    MTAG_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(M_dat, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$absn5, subset_data$value, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          MTAG_R_spearman = spearman_result$estimate,
                          MTAG_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    MTAG_mz_COR_results <- distinct(rbind(MTAG_mz_COR_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
MTAG_mz_COR_results$MTAG_adj <- p.adjust(MTAG_mz_COR_results$MTAG_p, method = "fdr")
MTAG_mz_COR_results<-MTAG_mz_COR_results%>%arrange(MTAG_p)
MTAG_mz_COR_results$sig_MTAG <- ifelse(MTAG_mz_COR_results$MTAG_adj < 0.01, "SIG", "NS")

sig_MTAG_mz<-MTAG_mz_COR_results%>%filter(sig_MTAG=="SIG")%>%pull(COMPOUND)
```
Mass Correlations across all metabolites for females
```{r}
F_mass<-melted_mass%>%filter(sex=="female")

# Create an empty results data frame to store the results
Fmass_mz_COR_results <- data.frame(COMPOUND = character(0), 
                                    Fmass_R_spearman = numeric(0),
                                    Fmass_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(F_mass, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$massn5, subset_data$value, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          Fmass_R_spearman = spearman_result$estimate,
                          Fmass_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    Fmass_mz_COR_results <- distinct(rbind(Fmass_mz_COR_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
Fmass_mz_COR_results$Fmass_adj <- p.adjust(Fmass_mz_COR_results$Fmass_p, method = "fdr")
Fmass_mz_COR_results<-Fmass_mz_COR_results%>%arrange(Fmass_p)
Fmass_mz_COR_results$sig_Fmass <- ifelse(Fmass_mz_COR_results$Fmass_adj < 0.01, "SIG", "NS")

sig_Fmass_mz<-Fmass_mz_COR_results%>%filter(sig_Fmass=="SIG")%>%pull(COMPOUND)
```
Mass Correlations across all metabolites for males
```{r}
M_mass<-melted_mass%>%filter(sex=="male")

# Create an empty results data frame to store the results
Mmass_mz_COR_results <- data.frame(COMPOUND = character(0), 
                                    Mmass_R_spearman = numeric(0),
                                    Mmass_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(M_mass, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$massn5, subset_data$value, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          Mmass_R_spearman = spearman_result$estimate,
                          Mmass_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    Mmass_mz_COR_results <- distinct(rbind(Mmass_mz_COR_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
Mmass_mz_COR_results$Mmass_adj <- p.adjust(Mmass_mz_COR_results$Mmass_p, method = "fdr")
Mmass_mz_COR_results<-Mmass_mz_COR_results%>%arrange(Mmass_p)
Mmass_mz_COR_results$sig_Mmass <- ifelse(Mmass_mz_COR_results$Mmass_adj < 0.01, "SIG", "NS")

sig_Mmass_mz<-Mmass_mz_COR_results%>%filter(sig_Mmass=="SIG")%>%pull(COMPOUND)
```
rTAG Correlations across all metabolites for females
```{r}
# Create an empty results data frame to store the results
FrTAG_mz_COR_results <- data.frame(COMPOUND = character(0), 
                                    FrTAG_R_spearman = numeric(0),
                                    FrTAG_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(F_mass, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$rTAG, subset_data$value, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          FrTAG_R_spearman = spearman_result$estimate,
                          FrTAG_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    FrTAG_mz_COR_results <- distinct(rbind(FrTAG_mz_COR_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
FrTAG_mz_COR_results$FrTAG_adj <- p.adjust(FrTAG_mz_COR_results$FrTAG_p, method = "fdr")
FrTAG_mz_COR_results<-FrTAG_mz_COR_results%>%arrange(FrTAG_p)
FrTAG_mz_COR_results$sig_FrTAG <- ifelse(FrTAG_mz_COR_results$FrTAG_adj < 0.01, "SIG", "NS")

sig_FrTAG_mz<-FrTAG_mz_COR_results%>%filter(sig_FrTAG=="SIG")%>%pull(COMPOUND)
```
rTAG Correlations across all metabolites for females
```{r}
# Create an empty results data frame to store the results
MrTAG_mz_COR_results <- data.frame(COMPOUND = character(0), 
                                    MrTAG_R_spearman = numeric(0),
                                    MrTAG_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(M_mass, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$rTAG, subset_data$value, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          MrTAG_R_spearman = spearman_result$estimate,
                          MrTAG_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    MrTAG_mz_COR_results <- distinct(rbind(MrTAG_mz_COR_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
MrTAG_mz_COR_results$MrTAG_adj <- p.adjust(MrTAG_mz_COR_results$MrTAG_p, method = "fdr")
MrTAG_mz_COR_results<-MrTAG_mz_COR_results%>%arrange(MrTAG_p)
MrTAG_mz_COR_results$sig_MrTAG <- ifelse(MrTAG_mz_COR_results$MrTAG_adj < 0.01, "SIG", "NS")

sig_MrTAG_mz<-MrTAG_mz_COR_results%>%filter(sig_MrTAG=="SIG")%>%pull(COMPOUND)
```
Combining Phenotype Results
```{r}
Female_Results_All_samples<-merge(FTAG_mz_COR_results,Fmass_mz_COR_results,by="COMPOUND")
Female_Results_All_samples<-merge(Female_Results_All_samples,FrTAG_mz_COR_results,by="COMPOUND")

Male_Results_All_samples<-merge(MTAG_mz_COR_results,Mmass_mz_COR_results,by="COMPOUND")
Male_Results_All_samples<-merge(Male_Results_All_samples,MrTAG_mz_COR_results,by="COMPOUND")

WithinSex_Results_All_samples<-merge(Female_Results_All_samples,Male_Results_All_samples,by="COMPOUND")
```
Saving out results for summary document
```{r}
write.csv(FTAG_mz_COR_results,"output/04_All_Sample_Ftag_mz_results.csv",row.names=F)
write.csv(Fmass_mz_COR_results,"output/04_All_Sample_Fmass_mz_results.csv",row.names=F)
write.csv(FrTAG_mz_COR_results,"output/04_All_Sample_Frtag_mz_results.csv",row.names=F)
write.csv(MTAG_mz_COR_results,"output/04_All_Sample_Mtag_mz_results.csv",row.names=F)
write.csv(Mmass_mz_COR_results,"output/04_All_Sample_Mmass_mz_results.csv",row.names=F)
write.csv(MrTAG_mz_COR_results,"output/04_All_Sample_Mrtag_mz_results.csv",row.names=F)
write.csv(WithinSex_Results_All_samples,"output/04_All_Sample_pheno_mz_results.csv",row.names=F)
```
Saving out r files
```{r}
save(file='output/Within_Sex_TAG_mz_COR_results.data', WithinSex_Results_All_samples, sig_FTAG_mz, sig_MTAG_mz, sig_Fmass_mz, sig_Mmass_mz, sig_FrTAG_mz, sig_MrTAG_mz, mz)
rm(list = ls())
```