---
title:   "Sex Variation in the DGRP Metabolome - Univariate means analysis"
output:   html_notebook
by:       R. Coig
updated:  5/7/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Univariate analysis, TAG-Metabolite level correlations (Sample Means, without sex as a covariate)
```{r,load required libraries}
library(tidyr)#Data manipulation
library(dplyr)#Data manipulation
library(smatr)#Computing relative TAG
library(reshape2)
```
Loading in metabolite means data 
```{r,loading in metabolome data}
rm(list = ls())
load('../data/processed/mz_means.data')
load('../data/processed/batch_corrected.data')

load('../data/processed/phenotype-TAG.data')
load('../data/processed/phenotype-mass.data')
```
TAG Correlations across all metabolites for female means
```{r}
#formatting df
mz_means_by_TAG<-merge(mz_means_by_sex_geno,TAG_sex_geno_long,by=c("sex","genotype"))
mz_means_by_FTAG<-mz_means_by_TAG%>%filter(sex=="female")

# Create an empty results data frame to store the results
FTAG_mz_COR_means_results <- data.frame(COMPOUND = character(0), 
                                    FTAG_R_spearman = numeric(0),
                                    FTAG_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(mz_means_by_FTAG, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$tTAG_4_mu_ss, subset_data$mean, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          FTAG_R_spearman = spearman_result$estimate,
                          FTAG_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    FTAG_mz_COR_means_results <- distinct(rbind(FTAG_mz_COR_means_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
FTAG_mz_COR_means_results$FTAG_adj <- p.adjust(FTAG_mz_COR_means_results$FTAG_p, method = "fdr")
FTAG_mz_COR_means_results<-FTAG_mz_COR_means_results%>%arrange(FTAG_p)
FTAG_mz_COR_means_results$sig_FTAG <- ifelse(FTAG_mz_COR_means_results$FTAG_adj < 0.05, "SIG", "NS")

sig_meanFTAG_mz<-FTAG_mz_COR_means_results%>%filter(sig_FTAG=="SIG")%>%pull(COMPOUND)
```
TAG Correlations across all metabolites for male means
```{r}
#formatting df
mz_means_by_MTAG<-mz_means_by_TAG%>%filter(sex=="male")

# Create an empty results data frame to store the results
MTAG_mz_COR_means_results <- data.frame(COMPOUND = character(0), 
                                    MTAG_R_spearman = numeric(0),
                                    MTAG_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(mz_means_by_MTAG, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$tTAG_4_mu_ss, subset_data$mean, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          MTAG_R_spearman = spearman_result$estimate,
                          MTAG_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    MTAG_mz_COR_means_results <- distinct(rbind(MTAG_mz_COR_means_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
MTAG_mz_COR_means_results$MTAG_adj <- p.adjust(MTAG_mz_COR_means_results$MTAG_p, method = "fdr")
MTAG_mz_COR_means_results<-MTAG_mz_COR_means_results%>%arrange(MTAG_p)
MTAG_mz_COR_means_results$sig_MTAG <- ifelse(MTAG_mz_COR_means_results$MTAG_adj < 0.05, "SIG", "NS")

sig_meanMTAG_mz<-MTAG_mz_COR_means_results%>%filter(sig_MTAG=="SIG")%>%pull(COMPOUND)
```
Mass Correlations across all metabolites for female means
```{r}
#formatting df
mz_means_by_mass<-merge(mz_means_by_sex_geno,mass_sex_geno_long,by=c("sex","genotype"))
#Removing sex/geno combos that do not have mass values
mz_means_by_mass<-mz_means_by_mass%>%filter(!is.na(mass_4_mu_ss))
#formatting df
mz_means_by_Fmass<-mz_means_by_mass%>%filter(sex=="female")

# Create an empty results data frame to store the results
Fmass_mz_COR_means_results <- data.frame(COMPOUND = character(0), 
                                    Fmass_R_spearman = numeric(0),
                                    Fmass_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(mz_means_by_Fmass, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$mass_4_mu_ss, subset_data$mean, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          Fmass_R_spearman = spearman_result$estimate,
                          Fmass_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    Fmass_mz_COR_means_results <- distinct(rbind(Fmass_mz_COR_means_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
Fmass_mz_COR_means_results$Fmass_adj <- p.adjust(Fmass_mz_COR_means_results$Fmass_p, method = "fdr")
Fmass_mz_COR_means_results<-Fmass_mz_COR_means_results%>%arrange(Fmass_p)
Fmass_mz_COR_means_results$sig_Fmass <- ifelse(Fmass_mz_COR_means_results$Fmass_adj < 0.01, "SIG", "NS")

sig_meanFmass_mz<-Fmass_mz_COR_means_results%>%filter(sig_Fmass=="SIG")%>%pull(COMPOUND)
```
Mass Correlations across all metabolites for male means
```{r}
mz_means_by_Mmass<-mz_means_by_mass%>%filter(sex=="male")

# Create an empty results data frame to store the results
Mmass_mz_COR_means_results <- data.frame(COMPOUND = character(0), 
                                    Mmass_R_spearman = numeric(0),
                                    Mmass_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(mz_means_by_Mmass, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$mass_4_mu_ss, subset_data$mean, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          Mmass_R_spearman = spearman_result$estimate,
                          Mmass_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    Mmass_mz_COR_means_results <- distinct(rbind(Mmass_mz_COR_means_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
Mmass_mz_COR_means_results$Mmass_adj <- p.adjust(Mmass_mz_COR_means_results$Mmass_p, method = "fdr")
Mmass_mz_COR_means_results<-Mmass_mz_COR_means_results%>%arrange(Mmass_p)
Mmass_mz_COR_means_results$sig_Mmass <- ifelse(Mmass_mz_COR_means_results$Mmass_adj < 0.01, "SIG", "NS")

sig_meanMmass_mz<-Mmass_mz_COR_means_results%>%filter(sig_Mmass=="SIG")%>%pull(COMPOUND)
```
Creating relative TAG as the residuals of TAG~Mass
```{r}
rTAG_sex_geno_long<-merge(mass_sex_geno_long,TAG_sex_geno_long,by=c("sex","genotype"))%>%select(sex,genotype,tTAG_4_mu_ss,mass_4_mu_ss)%>%filter(!is.na(mass_4_mu_ss))
rTAG_sex_geno_long<-rTAG_sex_geno_long%>%filter(!is.na(tTAG_4_mu_ss))

# Perform SMA regression
sma_model <- sma(rTAG_sex_geno_long$tTAG_4_mu_ss ~ rTAG_sex_geno_long$mass_4_mu_ss, na.rm=T)

# Extract the residuals
rTAG_sex_geno_long$rTAG_4_mu_ss <- residuals(sma_model)

mz_means_by_rTAG<-merge(mz_means_by_sex_geno,rTAG_sex_geno_long,by=c("sex","genotype"))

```
rTAG Correlations across all metabolites for female means
```{r}
mz_means_by_FrTAG<-mz_means_by_rTAG%>%filter(sex=="female")

# Create an empty results data frame to store the results
FrTAG_mz_COR_means_results <- data.frame(COMPOUND = character(0), 
                                    FrTAG_R_spearman = numeric(0),
                                    FrTAG_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(mz_means_by_FrTAG, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$rTAG_4_mu_ss, subset_data$mean, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          FrTAG_R_spearman = spearman_result$estimate,
                          FrTAG_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    FrTAG_mz_COR_means_results <- distinct(rbind(FrTAG_mz_COR_means_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
FrTAG_mz_COR_means_results$FrTAG_adj <- p.adjust(FrTAG_mz_COR_means_results$FrTAG_p, method = "fdr")
FrTAG_mz_COR_means_results<-FrTAG_mz_COR_means_results%>%arrange(FrTAG_p)
FrTAG_mz_COR_means_results$sig_FrTAG <- ifelse(FrTAG_mz_COR_means_results$FrTAG_adj < 0.01, "SIG", "NS")

sig_meanFrTAG_mz<-FrTAG_mz_COR_means_results%>%filter(sig_FrTAG=="SIG")%>%pull(COMPOUND)
```
rTAG Correlations across all metabolites for female means
```{r}
mz_means_by_MrTAG<-mz_means_by_rTAG%>%filter(sex=="male")

# Create an empty results data frame to store the results
MrTAG_mz_COR_means_results <- data.frame(COMPOUND = character(0), 
                                    MrTAG_R_spearman = numeric(0),
                                    MrTAG_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(mz_means_by_MrTAG, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$rTAG_4_mu_ss, subset_data$mean, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          MrTAG_R_spearman = spearman_result$estimate,
                          MrTAG_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    MrTAG_mz_COR_means_results <- distinct(rbind(MrTAG_mz_COR_means_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
MrTAG_mz_COR_means_results$MrTAG_adj <- p.adjust(MrTAG_mz_COR_means_results$MrTAG_p, method = "fdr")
MrTAG_mz_COR_means_results<-MrTAG_mz_COR_means_results%>%arrange(MrTAG_p)
MrTAG_mz_COR_means_results$sig_MrTAG <- ifelse(MrTAG_mz_COR_means_results$MrTAG_adj < 0.01, "SIG", "NS")

sig_meanMrTAG_mz<-MrTAG_mz_COR_means_results%>%filter(sig_MrTAG=="SIG")%>%pull(COMPOUND)
```
Combining Phenotype Results
```{r}
Female_Results_Means<-merge(FTAG_mz_COR_means_results,Fmass_mz_COR_means_results,by="COMPOUND")
Female_Results_Means<-merge(Female_Results_Means,FrTAG_mz_COR_means_results,by="COMPOUND")

Male_Results_Means<-merge(MTAG_mz_COR_means_results,Mmass_mz_COR_means_results,by="COMPOUND")
Male_Results_Means<-merge(Male_Results_Means,MrTAG_mz_COR_means_results,by="COMPOUND")

WithinSex_Results_Means<-merge(Female_Results_Means,Male_Results_Means,by="COMPOUND")
```
Saving out results for summary document
```{r}
write.csv(FTAG_mz_COR_means_results,"output/04_Mean_Ftag_mz_results.csv",row.names=F)
write.csv(Fmass_mz_COR_means_results,"output/04_Mean_Fmass_mz_results.csv",row.names=F)
write.csv(FrTAG_mz_COR_means_results,"output/04_Mean_Frtag_mz_results.csv",row.names=F)
write.csv(MTAG_mz_COR_means_results,"output/04_Mean_Mtag_mz_results.csv",row.names=F)
write.csv(Mmass_mz_COR_means_results,"output/04_Mean_Mmass_mz_results.csv",row.names=F)
write.csv(MrTAG_mz_COR_means_results,"output/04_Mean_Mrtag_mz_results.csv",row.names=F)
write.csv(WithinSex_Results_Means,"output/04_Mean_pheno_mz_results.csv",row.names=F)
```
Saving out r files
```{r}
save(file='../data/results/Within_Sex_TAG_mz_Means_COR_results.data', WithinSex_Results_Means, mz)
#rm(list = ls())
```