---
title:   "Sex Variation in the DGRP Metabolome - Univariate means analysis"
output:   html_notebook
by:       R. Coig
updated:  6/1/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Univariate analysis, TAG-Metabolite level correlations (Sample Means, without sex as a covariate)
```{r,load required libraries}
library(tidyr)#Data manipulation
library(dplyr)#Data manipulation
library(smatr)#Computing relative TAG
library(reshape2)
```
Loading in metabolite means data 
```{r,loading in metabolome data}
rm(list = ls())
load('../data/processed/mz_means.data')
load('../data/processed/batch_corrected.data')

load('../data/processed/phenotype-dimorphism.data')
```
deltaTAG Correlations across all metabolites for female means
```{r}
#formatting df
mz_means_by_deltaTAG<-merge(mz_means_by_sex_geno,dimorphism_summary,by=c("genotype"))
mz_Fmeans_by_deltaTAG<-mz_means_by_deltaTAG%>%filter(sex=="female")

# Create an empty results data frame to store the results
deltaTAG_mz_COR_Fmeans_results <- data.frame(COMPOUND = character(0), 
                                    FdeltaTAG_R_spearman = numeric(0),
                                    FdeltaTAG_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(mz_Fmeans_by_deltaTAG, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$diff_TAG, subset_data$mean, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          FdeltaTAG_R_spearman = spearman_result$estimate,
                          FdeltaTAG_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    deltaTAG_mz_COR_Fmeans_results <- distinct(rbind(deltaTAG_mz_COR_Fmeans_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
deltaTAG_mz_COR_Fmeans_results$FdeltaTAG_adj <- p.adjust(deltaTAG_mz_COR_Fmeans_results$FdeltaTAG_p, method = "fdr")
deltaTAG_mz_COR_Fmeans_results<-deltaTAG_mz_COR_Fmeans_results%>%arrange(FdeltaTAG_p)
deltaTAG_mz_COR_Fmeans_results$sig_FdeltaTAG <- ifelse(deltaTAG_mz_COR_Fmeans_results$FdeltaTAG_adj < 0.05, "SIG", "NS")

sig_meanFdeltaTAG_mz<-deltaTAG_mz_COR_Fmeans_results%>%filter(sig_FdeltaTAG=="SIG")%>%pull(COMPOUND)
```
deltaTAG Correlations across all metabolites for male means
```{r}
#formatting df
mz_means_by_MdeltaTAG<-mz_means_by_deltaTAG%>%filter(sex=="male")

# Create an empty results data frame to store the results
deltaTAG_mz_COR_Mmeans_results <- data.frame(COMPOUND = character(0), 
                                    MdeltaTAG_R_spearman = numeric(0),
                                    MdeltaTAG_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(mz_means_by_MdeltaTAG, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$diff_TAG, subset_data$mean, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          MdeltaTAG_R_spearman = spearman_result$estimate,
                          MdeltaTAG_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    deltaTAG_mz_COR_Mmeans_results <- distinct(rbind(deltaTAG_mz_COR_Mmeans_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
deltaTAG_mz_COR_Mmeans_results$MdeltaTAG_adj <- p.adjust(deltaTAG_mz_COR_Mmeans_results$MdeltaTAG_p, method = "fdr")
deltaTAG_mz_COR_Mmeans_results<-deltaTAG_mz_COR_Mmeans_results%>%arrange(MdeltaTAG_p)
deltaTAG_mz_COR_Mmeans_results$sig_MdeltaTAG <- ifelse(deltaTAG_mz_COR_Mmeans_results$MdeltaTAG_adj < 0.05, "SIG", "NS")

sig_meanMdeltaTAG_mz<-deltaTAG_mz_COR_Mmeans_results%>%filter(sig_MdeltaTAG=="SIG")%>%pull(COMPOUND)
```
deltaMass Correlations across all metabolites for female means
```{r}
#formatting df
mz_means_by_deltamass<-merge(mz_means_by_sex_geno,dimorphism_summary,by=c("genotype"))
#Removing genotypes combos that do not have delta mass values
mz_means_by_deltamass<-mz_means_by_deltamass%>%filter(!is.na(diff_mass))
#formatting df
mz_Fmeans_by_deltamass<-mz_means_by_deltamass%>%filter(sex=="female")

# Create an empty results data frame to store the results
deltamass_mz_COR_Fmeans_results <- data.frame(COMPOUND = character(0), 
                                    Fdeltamass_R_spearman = numeric(0),
                                    Fdeltamass_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(mz_Fmeans_by_deltamass, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$diff_mass, subset_data$mean, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          Fdeltamass_R_spearman = spearman_result$estimate,
                          Fdeltamass_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    deltamass_mz_COR_Fmeans_results <- distinct(rbind(deltamass_mz_COR_Fmeans_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
deltamass_mz_COR_Fmeans_results$Fdeltamass_adj <- p.adjust(deltamass_mz_COR_Fmeans_results$Fdeltamass_p, method = "fdr")
deltamass_mz_COR_Fmeans_results<-deltamass_mz_COR_Fmeans_results%>%arrange(Fdeltamass_p)
deltamass_mz_COR_Fmeans_results$sig_Fdeltamass <- ifelse(deltamass_mz_COR_Fmeans_results$Fdeltamass_adj < 0.01, "SIG", "NS")

sig_meanFdeltamass_mz<-deltamass_mz_COR_Fmeans_results%>%filter(sig_Fdeltamass=="SIG")%>%pull(COMPOUND)
```
deltaMass Correlations across all metabolites for male means
```{r}
mz_Mmeans_by_deltamass<-mz_means_by_deltamass%>%filter(sex=="male")

# Create an empty results data frame to store the results
deltamass_mz_COR_Mmeans_results <- data.frame(COMPOUND = character(0), 
                                    Mdeltamass_R_spearman = numeric(0),
                                    Mdeltamass_p = numeric(0))

# Loop through unique PCs
for (metabolite in mz) {
  subset_data <- subset(mz_Mmeans_by_deltamass, COMPOUND == metabolite)

  if (nrow(subset_data) > 0) {
    # Calculate Spearman correlation for TAG and each metabolite
    spearman_result <- cor.test(subset_data$diff_mass, subset_data$mean, 
                                method = "spearman")


    # Create a temporary data frame for each PC
    temp_df <- data.frame(COMPOUND = metabolite, 
                          Mdeltamass_R_spearman = spearman_result$estimate,
                          Mdeltamass_p = spearman_result$p.value)
    
    # Append the temporary data frame to the results
    deltamass_mz_COR_Mmeans_results <- distinct(rbind(deltamass_mz_COR_Mmeans_results, temp_df))
  }
}

# Adjust p-values for False Discovery Rate for T
deltamass_mz_COR_Mmeans_results$Mdeltamass_adj <- p.adjust(deltamass_mz_COR_Mmeans_results$Mdeltamass_p, method = "fdr")
deltamass_mz_COR_Mmeans_results<-deltamass_mz_COR_Mmeans_results%>%arrange(Mdeltamass_p)
deltamass_mz_COR_Mmeans_results$sig_Mdeltamass <- ifelse(deltamass_mz_COR_Mmeans_results$Mdeltamass_adj < 0.01, "SIG", "NS")

sig_meanMmass_mz<-deltamass_mz_COR_Mmeans_results%>%filter(sig_Mdeltamass=="SIG")%>%pull(COMPOUND)
```
Combining delta Phenotype Results
```{r}
Female_Results_Means<-merge(deltaTAG_mz_COR_Fmeans_results,deltamass_mz_COR_Fmeans_results,by="COMPOUND")

Male_Results_Means<-merge(deltaTAG_mz_COR_Mmeans_results,deltamass_mz_COR_Mmeans_results,by="COMPOUND")

WithinSex_delta_Results_Means<-merge(Female_Results_Means,Male_Results_Means,by="COMPOUND")
```
Saving out results for summary document
```{r}
write.csv(deltaTAG_mz_COR_Fmeans_results,"output/04b_Mean_FdeltaTAG_mz_results.csv",row.names=F)
write.csv(deltamass_mz_COR_Fmeans_results,"output/04b_Mean_Fdeltamass_mz_results.csv",row.names=F)
write.csv(deltaTAG_mz_COR_Mmeans_results,"output/04b_Mean_MdeltaTAG_mz_results.csv",row.names=F)
write.csv(deltamass_mz_COR_Mmeans_results,"output/04b_Mean_Mmass_mz_results.csv",row.names=F)
write.csv(WithinSex_delta_Results_Means,"output/04b_deltaTAG_withinsex_mz_results.csv",row.names=F)
```
Saving out r files
```{r}
save(file='../data/results/Within_Sex_deltaTAG_mz_Means_COR_results.data', WithinSex_delta_Results_Means, mz)
#rm(list = ls())
```