---
title:   "Sex Variation in the DGRP Metabolome - SD Correlations"
output:   html_notebook
by:       R. Coig
updated:  6/1/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Metabolite Correlations across all samples
```{r,load required libraries}
library(reshape2)
library(tidyr)
library(dplyr)
```
Loading in metabolome data for analysis
```{r,loading in metabolome data}
rm(list = ls())
load('../data/processed/batch_corrected.data')
source("../src/compute_p_values_matrix.R")
source("../src/remove_duplicate_cors.R")
```
Create correlation matrix across all metabolites and all samples using Spearman's correlation (non-parametric)
```{r}
# Calculate correlations for all metabolites
correlation_matrix <- cor(dat_by_sample, method = "spearman")
correlation_matrix<-as.matrix(correlation_matrix)

# Create an empty matrix to store adjusted p-values
p_matrix <- matrix(NA, nrow = ncol(correlation_matrix), ncol = ncol(correlation_matrix))

# Loop through each pair of columns and compute adjusted p-values
for (i in 1:ncol(correlation_matrix)) {
  for (j in 1:ncol(correlation_matrix)) {
    p_matrix[i, j] <- compute_p_values(correlation_matrix[, i], correlation_matrix[, j])
  }
}
for (i in 1:ncol(correlation_matrix)) {
  for (j in 1:ncol(correlation_matrix)) {
    p_matrix[i, j] <- compute_p_values(correlation_matrix[, i], correlation_matrix[, j])
  }
}

# Adjust the p-values matrix
adjusted_matrix <- p.adjust(p_matrix, method = "fdr")
# Ensure that the adjusted p-values matrix remains a matrix
dim(adjusted_matrix) <- dim(p_matrix)

rownames(adjusted_matrix)<-rownames(correlation_matrix)
colnames(adjusted_matrix)<-colnames(correlation_matrix)
```
correlations FDR <.05
```{r}
# Initialize an empty data frame to store the significant correlations
significant_correlations <- data.frame(rowname = character(),
                                          colname = character(),
                                          pvalue = numeric(),
                                          stringsAsFactors = FALSE)

# Loop through the indices of the adjusted p-value matrix
for (i in 1:nrow(adjusted_matrix)) {
  for (j in 1:ncol(adjusted_matrix)) {
    if (adjusted_matrix[i, j] < 0.01 && i != j) {  # Check if rowname and colname are different
      # Store the significant correlation information in the data frame
      significant_correlations[nrow(significant_correlations) + 1,] <- list(rownames(adjusted_matrix)[i],
                                                                                   colnames(adjusted_matrix)[j],
                                                                                   adjusted_matrix[i, j])
    }
  }
}

# Remove rows where A,B = B,A
significant_correlations <- remove_duplicates(significant_correlations)%>%arrange(pvalue)
```
Adding Spearman's rho
```{r}
# Initialize an empty data frame to store the significant correlations
correlations_rho <- data.frame(rowname = character(),
                                          colname = character(),
                                          R_squared = numeric(),
                                          stringsAsFactors = FALSE)

# Loop through the indices of the adjusted p-value matrix
for (i in 1:nrow(correlation_matrix)) {
  for (j in 1:ncol(correlation_matrix)) {
    if (i != j) {  # Check if rowname and colname are different
      # Store the significant correlation information in the data frame
      correlations_rho[nrow(correlations_rho) + 1,] <- list(rownames(correlation_matrix)[i],
                                                                                   colnames(correlation_matrix)[j],
                                                                                   correlation_matrix[i, j])
    }
  }
}

Correlations<-merge(significant_correlations,correlations_rho,by=c("rowname","colname"))
```
With rho greater than .80
```{r}
strong_anticorrelations<-Correlations%>%filter(R_squared < -0.80)%>%arrange(R_squared)
strong_correlations<-Correlations%>%dplyr::filter(R_squared > 0.80)%>%arrange(-R_squared)
```
Saving out correlations and p values csvs
```{r}
write.csv(Correlations,"output/06a_Allsample_correlations_results.csv",row.names=T)
write.csv(correlation_matrix,"output/06a_Allsample_correlation_matrix.csv",row.names=T)
write.csv(adjusted_matrix,"output/06a_Allsample_adjpvalues_matrix.csv",row.names=T)
write.csv(strong_anticorrelations,"output/06a_Allsample_strongest_anticorrelations.csv",row.names=T)
write.csv(strong_correlations,"output/06a_Allsample_strongest_correlations.csv",row.names=T)
```
Saving out .rds files
```{r,save out r files}
save(file='../data/results/Correlations_results.data', Correlations, mz)
```