---
title:   "Sex Variation in the DGRP Metabolome - SD Correlations"
output:   html_notebook
by:       R. Coig
updated:  6/1/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Metabolite SD Correlations across all samples
```{r,load required libraries}
library(reshape2)
library(tidyr)
library(dplyr)
```
Loading in metabolome data for analysis
```{r,loading in metabolome data}
rm(list = ls())
load('../data/processed/batch_corrected.data')
source("../src/compute_p_values_matrix.R")
source("../src/remove_duplicate_cors.R")
```
Create correlation matrix of sex differences across all metabolites using Spearman's correlation (non-parametric)
```{r}
# Extract columns 15 to 74
female_samples<-dat_key%>%filter(sex=="female")
male_samples<-dat_key%>%filter(sex=="male")
selected_females <- female_samples[, 11:195]
selected_males <- male_samples[, 11:195]

# Calculate correlations for the selected columns
correlation_matrix_F <- cor(selected_females, method = "spearman")
correlation_matrix_M <- cor(selected_males, method = "spearman")

correlation_matrix_F<-as.matrix(correlation_matrix_F)
correlation_matrix_M<-as.matrix(correlation_matrix_M)

# Create an empty matrix to store adjusted p-values
p_F_matrix <- matrix(NA, nrow = ncol(correlation_matrix_F), ncol = ncol(correlation_matrix_F))
p_M_matrix <- matrix(NA, nrow = ncol(correlation_matrix_M), ncol = ncol(correlation_matrix_M))

# Loop through each pair of columns and compute adjusted p-values
for (i in 1:ncol(correlation_matrix_F)) {
  for (j in 1:ncol(correlation_matrix_F)) {
    p_F_matrix[i, j] <- compute_p_values(correlation_matrix_F[, i], correlation_matrix_F[, j])
  }
}
for (i in 1:ncol(correlation_matrix_M)) {
  for (j in 1:ncol(correlation_matrix_M)) {
    p_M_matrix[i, j] <- compute_p_values(correlation_matrix_M[, i], correlation_matrix_M[, j])
  }
}

# Adjust the p-values matrix
adjusted_F_matrix <- p.adjust(p_F_matrix, method = "fdr")
adjusted_M_matrix <- p.adjust(p_M_matrix, method = "fdr")
# Ensure that the adjusted p-values matrix remains a matrix
dim(adjusted_F_matrix) <- dim(p_F_matrix)
dim(adjusted_M_matrix) <- dim(p_M_matrix)

rownames(adjusted_F_matrix)<-rownames(correlation_matrix_F)
colnames(adjusted_F_matrix)<-colnames(correlation_matrix_F)
rownames(adjusted_M_matrix)<-rownames(correlation_matrix_M)
colnames(adjusted_M_matrix)<-colnames(correlation_matrix_M)


```
Within-sex correlations FDR <.05
```{r}
# Initialize an empty data frame to store the significant correlations
significant_correlations_F <- data.frame(rowname = character(),
                                          colname = character(),
                                          pvalue = numeric(),
                                          stringsAsFactors = FALSE)
significant_correlations_M <- data.frame(rowname = character(),
                                          colname = character(),
                                          pvalue = numeric(),
                                          stringsAsFactors = FALSE)

# Loop through the indices of the adjusted p-value matrix
for (i in 1:nrow(adjusted_F_matrix)) {
  for (j in 1:ncol(adjusted_F_matrix)) {
    if (adjusted_F_matrix[i, j] < 0.05 && i != j) {  # Check if rowname and colname are different
      # Store the significant correlation information in the data frame
      significant_correlations_F[nrow(significant_correlations_F) + 1,] <- list(rownames(adjusted_F_matrix)[i],
                                                                                   colnames(adjusted_F_matrix)[j],
                                                                                   adjusted_F_matrix[i, j])
    }
  }
}

for (i in 1:nrow(adjusted_M_matrix)) {
  for (j in 1:ncol(adjusted_M_matrix)) {
    if (adjusted_M_matrix[i, j] < 0.05 && i != j) {  # Check if rowname and colname are different
      # Store the significant correlation information in the data frame
      significant_correlations_M[nrow(significant_correlations_M) + 1,] <- list(rownames(adjusted_M_matrix)[i],
                                                                                   colnames(adjusted_M_matrix)[j],
                                                                                   adjusted_M_matrix[i, j])
    }
  }
}

# Remove rows where A,B = B,A
significant_correlations_F <- remove_duplicates(significant_correlations_F)%>%arrange(pvalue)
significant_correlations_M <- remove_duplicates(significant_correlations_M)%>%arrange(pvalue)
```
Adding Spearman's Rho
```{r}
# Initialize an empty data frame to store the significant correlations
Correlations_F <- data.frame(rowname = character(),
                                          colname = character(),
                                          R_squared = numeric(),
                                          stringsAsFactors = FALSE)
Correlations_M <- data.frame(rowname = character(),
                                          colname = character(),
                                          R_squared = numeric(),
                                          stringsAsFactors = FALSE)

# Loop through the indices of the adjusted p-value matrix
for (i in 1:nrow(correlation_matrix_F)) {
  for (j in 1:ncol(correlation_matrix_F)) {
    if (i != j) {  # Check if rowname and colname are different
      # Store the significant correlation information in the data frame
      Correlations_F[nrow(Correlations_F) + 1,] <- list(rownames(correlation_matrix_F)[i],
                                                                                   colnames(correlation_matrix_F)[j],
                                                                                   correlation_matrix_F[i, j])
    }
  }
}
for (i in 1:nrow(correlation_matrix_M)) {
  for (j in 1:ncol(correlation_matrix_M)) {
    if (i != j) {  # Check if rowname and colname are different
      # Store the significant correlation information in the data frame
      Correlations_M[nrow(Correlations_M) + 1,] <- list(rownames(correlation_matrix_M)[i],
                                                                                   colnames(correlation_matrix_M)[j],
                                                                                   correlation_matrix_M[i, j])
    }
  }
}
Correlations_F_allsamples<-merge(significant_correlations_F,Correlations_F,by=c("rowname","colname"))%>%arrange(pvalue)%>%dplyr::rename(p_F_all=pvalue,R_F_all=R_squared)
Correlations_M_allsamples<-merge(significant_correlations_M,Correlations_M,by=c("rowname","colname"))%>%arrange(pvalue)%>%dplyr::rename(p_M_all=pvalue,R_M_all=R_squared)
```
With rho greater than .80
```{r}
strong_anticorrelations_F_allsamples<-Correlations_F_allsamples%>%filter(R_F_all < -0.80)%>%arrange(R_F_all)
strong_correlations_F_allsamples<-Correlations_F_allsamples%>%dplyr::filter(R_F_all > 0.80)%>%arrange(-R_F_all)

strong_anticorrelations_M_allsamples<-Correlations_M_allsamples%>%filter(R_M_all < -0.80)%>%arrange(R_M_all)
strong_correlations_M_allsamples<-Correlations_M_allsamples%>%dplyr::filter(R_M_all > 0.80)%>%arrange(-R_M_all)
```
Saving out correlations and p values csvs
```{r}
write.csv(Correlations_M_allsamples,"output/06a_Male_correlations_results.csv",row.names=T)
write.csv(strong_correlations_M_allsamples,"output/06a_Male_strongest_correlations_results.csv",row.names=T)
write.csv(strong_anticorrelations_M_allsamples,"output/06a_Male_strongest_anticorrelations_results.csv",row.names=T)
write.csv(Correlations_F_allsamples,"output/06a_Female_correlations_results.csv",row.names=T)
write.csv(strong_correlations_F_allsamples,"output/06a_Female_strongest_correlations_results.csv",row.names=T)
write.csv(strong_anticorrelations_F_allsamples,"output/06a_Female_strongest_anticorrelations_results.csv",row.names=T)
```
Saving out .rds files
```{r,save out r files}
save(file='../data/results/WithinSex_Allsamples_Correlated_mzs_results.data', Correlations_F_allsamples, Correlations_M_allsamples, mz)
```