---
title:   "Sex Variation in the DGRP Metabolome - SD Correlations"
output:   html_notebook
by:       R. Coig
updated:  6/1/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Metabolite Correlations across sex/geno means
```{r,load required libraries}
library(reshape2)
library(tidyr)
library(dplyr)
```
Loading in metabolome data for analysis
```{r,loading in metabolome data}
rm(list = ls())
load('../data/processed/mz_means.data')
source("../src/compute_p_values_matrix.R")
source("../src/remove_duplicate_cors.R")
```
Create correlation matrix of sex differences across all metabolites using Spearman's correlation (non-parametric)
```{r}
# Pivot the data from long to wide format
wide_mz_means <- mz_means_by_sex_geno[,1:4] %>%
  pivot_wider(names_from = COMPOUND, values_from = mean)

wide_mz_means<-wide_mz_means[,-c(1:2)]

# Calculate correlations for the selected columns
correlation_matrix_means <- cor(wide_mz_means, method = "spearman")
correlation_matrix_means<-as.matrix(correlation_matrix_means)

# Create an empty matrix to store adjusted p-values
p_means_matrix <- matrix(NA, nrow = ncol(correlation_matrix_means), ncol = ncol(correlation_matrix_means))

# Loop through each pair of columns and compute adjusted p-values
for (i in 1:ncol(correlation_matrix_means)) {
  for (j in 1:ncol(correlation_matrix_means)) {
    p_means_matrix[i, j] <- compute_p_values(correlation_matrix_means[, i], correlation_matrix_means[, j])
  }
}

# Adjust the p-values matrix
adjusted_matrix <- p.adjust(p_means_matrix, method = "fdr")
# Ensure that the adjusted p-values matrix remains a matrix
dim(adjusted_matrix) <- dim(p_means_matrix)

rownames(adjusted_matrix)<-rownames(correlation_matrix_means)
colnames(adjusted_matrix)<-colnames(correlation_matrix_means)
```
Sex/Geno means correlations FDR <.05
```{r}
# Initialize an empty data frame to store the significant correlations
significant_correlations_means <- data.frame(rowname = character(),
                                          colname = character(),
                                          pvalue = numeric(),
                                          stringsAsFactors = FALSE)

# Loop through the indices of the adjusted p-value matrix
for (i in 1:nrow(adjusted_matrix)) {
  for (j in 1:ncol(adjusted_matrix)) {
    if (adjusted_matrix[i, j] < 0.05 && i != j) {  # Check if rowname and colname are different
      # Store the significant correlation information in the data frame
      significant_correlations_means[nrow(significant_correlations_means) + 1,] <- list(rownames(adjusted_matrix)[i],
                                                                                   colnames(adjusted_matrix)[j],
                                                                                   adjusted_matrix[i, j])
    }
  }
}

# Remove rows where A,B = B,A
significant_correlations_means <- remove_duplicates(significant_correlations_means)%>%arrange(pvalue)
```
Adding Spearman's Rho
```{r}
# Initialize an empty data frame to store the significant correlations
Correlations_Means <- data.frame(rowname = character(),
                                          colname = character(),
                                          R_squared = numeric(),
                                          stringsAsFactors = FALSE)

# Loop through the indices of the adjusted p-value matrix
for (i in 1:nrow(correlation_matrix_means)) {
  for (j in 1:ncol(correlation_matrix_means)) {
    if (i != j) {  # Check if rowname and colname are different
      # Store the significant correlation information in the data frame
      Correlations_Means[nrow(Correlations_Means) + 1,] <- list(rownames(correlation_matrix_means)[i],
                                                                                   colnames(correlation_matrix_means)[j],
                                                                                   correlation_matrix_means[i, j])
    }
  }
}

Correlations_Means<-merge(significant_correlations_means,Correlations_Means,by=c("rowname","colname"))%>%arrange(pvalue)%>%dplyr::rename(p_means=pvalue,R_means=R_squared)

```
With rho greater than .80
```{r}
strong_anticorrelations_means<-Correlations_Means%>%filter(R_means < -0.80)%>%arrange(R_means)
strong_correlations_means<-Correlations_Means%>%dplyr::filter(R_means > 0.80)%>%arrange(-R_means)
```
Saving out correlations and p values csvs
```{r}
write.csv(Correlations_Means,"output/06b_Sexgenomeans_correlations_results.csv",row.names=T)
write.csv(strong_correlations_means,"output/06b_sexgenomeans_strongest_correlations_results.csv",row.names=T)
write.csv(strong_anticorrelations_means,"output/06b_sexgenomeans_strongest_anticorrelations_results.csv",row.names=T)

```
Saving out .rds files
```{r,save out r files}
save(file='../data/results/SexGenomeans_Correlated_mzs_results.data', Correlations_Means, mz)
```