---
title:   "Sex Variation in the DGRP Metabolome - SD Correlations"
output:   html_notebook
by:       R. Coig
updated:  6/1/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Metabolite SD Correlations across all samples
```{r,load required libraries}
library(reshape2)
library(tidyr)
library(dplyr)
```
Loading in metabolome data for analysis
```{r,loading in metabolome data}
rm(list = ls())
load('../data/processed/mz_SD.data')
source("../src/compute_p_values_matrix.R")
source("../src/remove_duplicate_cors.R")
```
Create correlation matrix of sex differences across all metabolites using Spearman's correlation (non-parametric)
```{r}
# Pivot the data from long to wide format
wide_geno_diffs <- SD_by_geno[,1:3] %>%
  pivot_wider(names_from = COMPOUND, values_from = mz_SD)

# Extract columns 15 to 74
selected_columns <- wide_geno_diffs[, 2:186]

# Calculate correlations for the selected columns
correlation_matrix_SD <- cor(selected_columns, method = "spearman")

selected_columns_matrix<-as.matrix(selected_columns)

# Create an empty matrix to store adjusted p-values
p_value_matrix <- matrix(NA, nrow = ncol(selected_columns_matrix), ncol = ncol(selected_columns_matrix))

# Loop through each pair of columns and compute adjusted p-values
for (i in 1:ncol(selected_columns_matrix)) {
  for (j in 1:ncol(selected_columns_matrix)) {
    p_value_matrix[i, j] <- compute_p_values(selected_columns_matrix[, i], selected_columns_matrix[, j])
  }
}

# Adjust the p-values matrix
adjusted_SD_matrix <- p.adjust(p_value_matrix, method = "fdr")
# Ensure that the adjusted p-values matrix remains a matrix
dim(adjusted_SD_matrix) <- dim(p_value_matrix)
rownames(adjusted_SD_matrix)<-rownames(correlation_matrix_SD)
colnames(adjusted_SD_matrix)<-colnames(correlation_matrix_SD)
```
SD correlations FDR <.05
```{r}
# Initialize an empty data frame to store the significant correlations
significant_correlations_SD <- data.frame(rowname = character(),
                                          colname = character(),
                                          pvalue = numeric(),
                                          stringsAsFactors = FALSE)

# Loop through the indices of the adjusted p-value matrix
for (i in 1:nrow(adjusted_SD_matrix)) {
  for (j in 1:ncol(adjusted_SD_matrix)) {
    if (adjusted_SD_matrix[i, j] < 0.05 && i != j) {  # Check if rowname and colname are different
      # Store the significant correlation information in the data frame
      significant_correlations_SD[nrow(significant_correlations_SD) + 1,] <- list(rownames(adjusted_SD_matrix)[i],
                                                                                   colnames(adjusted_SD_matrix)[j],
                                                                                   adjusted_SD_matrix[i, j])
    }
  }
}
# Remove rows where A,B = B,A
significant_correlations_SD <- remove_duplicates(significant_correlations_SD)%>%arrange(pvalue)%>%dplyr::rename(p_SD=pvalue)
```
Adding Spearman's Rho
```{r}
# Initialize an empty data frame to store the significant correlations
Correlations_SD <- data.frame(rowname = character(),
                                          colname = character(),
                                          R_SD = numeric(),
                                          stringsAsFactors = FALSE)

# Loop through the indices of the adjusted p-value matrix
for (i in 1:nrow(correlation_matrix_SD)) {
  for (j in 1:ncol(correlation_matrix_SD)) {
    if (i != j) {  # Check if rowname and colname are different
      # Store the significant correlation information in the data frame
      Correlations_SD[nrow(Correlations_SD) + 1,] <- list(rownames(correlation_matrix_SD)[i],
                                                                                   colnames(correlation_matrix_SD)[j],
                                                                                   correlation_matrix_SD[i, j])
    }
  }
}
Correlations_SD<-merge(significant_correlations_SD,Correlations_SD,by=c("rowname","colname"))
```

With rho greater than .80
```{r}
strong_anticorrelations_SD<-Correlations_SD%>%filter(R_SD < -0.8)%>%arrange(R_SD)
strong_correlations_SD<-Correlations_SD%>%dplyr::filter(R_SD > 0.8)%>%arrange(-R_SD)
```
Saving out correlations and p values csvs
```{r}
write.csv(Correlations_SD,"output/06c_sex_diff_correlations_results.csv",row.names=T)
write.csv(strong_anticorrelations_SD,"output/06c_strong_anticorrelations_SD.csv",row.names=F)
write.csv(strong_correlations_SD,"output/06c_strong_correlations_SD.csv",row.names=F)
```
Saving out .rds files
```{r,save out r files}
save(file='../data/results/Correlated_SDmzs_results.data', Correlations_SD, mz)
```