---
title:   "Sex Variation in the DGRP Metabolome - Phenotype"
output:   html_notebook
by:       R. Coig
updated:  3/31/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Phenotype summaries for mass 
---
Load required packages.
```{r}
library(tidyr)#data manipulation
library(dplyr)#data manipulation
library(reshape2)#data manipulation
```
Reading in raw mass data. Masses were measured as 5 flies per sample.
```{r}
rm(list = ls())
mass <- read.csv("data/raw/MassSamples.csv")#raw mass data
mass$genotype<-as.factor(mass$genotype)
mass$sex<-as.factor(mass$sex)
```
Calculating mass summaries
```{r}
B7_mass_summary_sex_geno <- mass %>%
  dplyr::group_by(sex, genotype) %>%
  dplyr::reframe(
    genotype = first(genotype),
    sex = first(sex),
    mass_4_mu_ss = ifelse(all(is.na(massn5)), NA, round(mean(massn5, na.rm = TRUE), 2)),
    mass_4_sd_ss = round(sd(massn5,na.rm=T), 2),
    mass_4_var_ss = round(var(massn5,na.rm=T), 2),  # Calculate variance
    mass_4_cv_ss = round((mass_4_sd_ss / mass_4_mu_ss) * 100, 2),  # Calculate CV
    mass_4_se_ss = round(mass_4_sd_ss / sqrt(n()), 2),  # Calculate SE
    mass_4_n_ss = n()) %>% # Calculate sample size
  distinct()

B7_mass_summary_sex_geno_wide <- B7_mass_summary_sex_geno %>%
  pivot_wider(
    id_cols = genotype,
    names_from = sex,
    values_from = c(mass_4_mu_ss, mass_4_sd_ss, mass_4_var_ss, mass_4_cv_ss, mass_4_se_ss, mass_4_n_ss)
  )

B7_mass_summary_geno <- mass %>%
  dplyr::group_by(genotype) %>%
  dplyr::summarize(
    genotype = first(genotype),
    mass_4_mu_line = ifelse(all(is.na(massn5)), NA, round(mean(massn5, na.rm = TRUE), 2)),
    mass_4_sd_line = round(sd(massn5,na.rm=T), 2),
    mass_4_var_line = round(var(massn5,na.rm=T), 2),  # Calculate variance
    mass_4_cv_line = round((mass_4_sd_line / mass_4_mu_line) * 100, 2),  # Calculate CV
    mass_4_se_line = round(mass_4_sd_line / sqrt(n()), 2),  # Calculate SE
    mass_4_n_line = n())

B7_mass_summary_sex <- mass %>%
  dplyr::group_by(sex) %>%
  dplyr::summarize(
    sex = first(sex),
    mass_4_mu = ifelse(all(is.na(massn5)), NA, round(mean(massn5, na.rm = TRUE), 2)),
    mass_4_sd = round(sd(massn5,na.rm=T), 2),
    mass_4_var = round(var(massn5,na.rm=T), 2),  # Calculate variance
    mass_4_cv = round((mass_4_sd / mass_4_mu) * 100, 2),  # Calculate CV
    mass_4_se = round(mass_4_sd / sqrt(n()), 2),  # Calculate SE
    mass_4_n = n())
    
B7_mass_pooled <- mass %>%
  filter(!is.na(massn5)) %>%
  dplyr::group_by(genotype) %>%
  dplyr::summarise(
    n_female = sum(sex == "female"),
    n_male = sum(sex == "male"),
    sd_female = sd(massn5[sex == "female"]),
    sd_male = sd(massn5[sex == "male"]),
    mass_4_pl_sd = round(sqrt(((n_female - 1) * sd_female^2 + (n_male - 1) * sd_male^2) / (n_female + n_male - 2)),4),  # Calculate pooled_sd between sexes
    mass_4_pl_se = round(sqrt(((n_female - 1) * sd_female^2 + (n_male - 1) * sd_male^2) / (n_female + n_male - 2)) / sqrt(n_female + n_male),4)  # Calculate pooled_se between sexes
  ) %>%
  select(genotype, mass_4_pl_sd, mass_4_pl_se)
```
Combining summaries into dfs for later analysis.
```{r}
mass_sex_geno_wide<- left_join(B7_mass_summary_geno,B7_mass_summary_sex_geno_wide,by="genotype")
mass_sex_geno_wide<- left_join(mass_sex_geno_wide,B7_mass_pooled,by="genotype")
mass_sex_geno_long<- left_join(B7_mass_summary_sex_geno,B7_mass_summary_geno, by="genotype")
```
Saving out summaries 
```{r}
write.csv(mass_sex_geno_wide,"output/01_mass_summary_wide_format.csv",row.names=F)
write.csv(mass_sex_geno_long,"output/01_mass_summary_long_format.csv",row.names=F)
write.csv(B7_mass_summary_sex,"output/01_mass_population_sex_summary.csv",row.names=F)
```
saving out r data files
```{r}
save(file='../data/processed/phenotype-mass.data', mass_sex_geno_wide, mass_sex_geno_long)
rm(list=ls())
```