---
title:   "Sex Variation in the DGRP Metabolome - Phenotype"
output:   html_notebook
by:       R. Coig
updated:  3/31/24
---
Project Summary: The overarching goal of this project is to investigate the metabolome of flies that are highly dimorphic and non dimorphic for body fat and mass phenotypes, to identify molecular mechanisms driving continuous variation in sexual size dimorphism. These metabolome data come from 141 head samples of males and females from 29 DGRP lines (Mackay et al, 2012) previously identified as "high dimorphism" or "low dimorphism" for body fat. 

Notebook Summary: Calculating Dimorphism metrics for TAG and mass
---
Load required packages.
```{r}
library(tidyr)#data manipulation
library(dplyr)#data manipulation
library(reshape2)#data manipulation
```
Reading in raw phenotype data for TAG and mass
```{r}
rm(list = ls())
load('../data/processed/phenotype-TAG.data')
load('../data/processed/phenotype-mass.data')
```
Calculating the values of high and low dimorphism for the genotype-specific mean values of all 3 phenotypes using two metrics:
1) F-M subtracting male from female mean values by genotype (diff_tTAG)
2) log (F/M) (log_div_tTAG)
```{r}
diff_mass <- mass_sex_geno_wide[,c(1,8,9)]
diff_mass$diff_mass <- diff_mass$mass_4_mu_ss_female-diff_mass$mass_4_mu_ss_male
diff_mass$log_div_mass =round(log(diff_mass$mass_4_mu_ss_female/diff_mass$mass_4_mu_ss_male),2)

diff_TAG <- TAG_sex_geno_wide[,c(1,8,9)]
diff_TAG$diff_TAG <- diff_TAG$tTAG_4_mu_ss_female-diff_TAG$tTAG_4_mu_ss_male
diff_TAG$log_div_TAG =round(log(diff_TAG$tTAG_4_mu_ss_female/diff_TAG$tTAG_4_mu_ss_male),4)

dimorphism_summary<-left_join(diff_TAG,diff_mass,by="genotype")
```
Summary SSD stats
```{r}
SSD_summary <- dimorphism_summary %>%
  dplyr::summarize(
    SSD_TAG_mu = round(mean(diff_TAG), 4),
    SSD_TAG_sd = round(sd(diff_TAG), 4),
    SSD_TAG_var = round(var(diff_TAG), 4),  # Calculate variance
    SSD_TAG_cv = round((SSD_TAG_sd / SSD_TAG_mu) * 100, 4),  # Calculate CV
    SSD_TAG_se = round(SSD_TAG_sd / sqrt(n()), 4),  # Calculate SE
    SSD_TAG_n = n(),
    SSD_mass_mu = round(mean(diff_mass,na.rm=T), 2),
    SSD_mass_sd = round(sd(diff_mass,na.rm=T), 2),
    SSD_mass_var = round(var(diff_mass,na.rm=T), 2),  # Calculate variance
    SSD_mass_cv = round((SSD_mass_sd / SSD_mass_mu) * 100, 2),  # Calculate CV
    SSD_mass_se = round(SSD_mass_sd / sqrt(n()), 2),  # Calculate SE
    SSD_mass_n = n()    )
```
Save out results:
```{r,save out results in csv format}
write.csv(dimorphism_summary,"output/01_dimorphism_summary.csv",row.names=F)
write.csv(SSD_summary,"output/01_SSD_pop_summary.csv",row.names=F)
```
saving out r data files
```{r}
save(file='../data/processed/phenotype-dimorphism.data', dimorphism_summary)
rm(list=ls())
```